### 2024-02-12 paper/paper.qmd
Started replicating fig 1

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Aaron Liu
  - Jimmy Luc
  - Maroosh Gillani
thanks: "Code and data are available at: https://github.com/rex009x/us_birth_rates/tree/master/data . A replication of various aspects in this paper are available at: LINK ."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---
```{r, include=FALSE, warning=FALSE, message=FALSE, echo=FALSE}
# workplace setup

library(tidyverse)
library(dplyr)
library(ggplot2)
library(here)

# acquire fig_1_data from gig_1.csv
fig_1_data <- 
  read.csv(file=here("data/fig_1.csv"))
```

# Introduction

# CHANGE THIS

```{r graph1, fig.cap="Trend in US Birth Rates from 1980 - 2020", echo=FALSE}

#| label: fig-1

# replicate figure 1 from the original paper
fig_1_data %>% ggplot(aes(x = year, y = brate_all)) +
  geom_line(color = "blue") +
  labs(x = "Year", y = "Births per 1,000 women age 15–44")
```

# References

### 2024-02-12 paper/paper.qmd
Figure 1 basically replicated

@@
 author: 
+  - Maroosh Gillani
   - Aaron Liu
   - Jimmy Luc
-  - Maroosh Gillani
 thanks: "Code and data are available at: https://github.com/rex009x/us_birth_rates/tree/master/data . A replication of various aspects in this paper are available at: LINK ."
@@
 fig_1_data %>% ggplot(aes(x = year, y = brate_all)) +
-  geom_line(color = "blue") +
-  labs(x = "Year", y = "Births per 1,000 women age 15–44")
+  geom_line(color = "#378bd4", size = 1) +
+  labs(x = "Year", y = "Births per 1,000 women age 15–44") +
+  scale_y_continuous(limits = c(50,80), breaks = seq(50, 80, by = 5)) +
+  scale_x_continuous(limits = c(1980,2020), breaks = seq(1980, 2020, by = 5)) +
+  theme_classic()
 ```

### 2024-02-12 paper/paper.qmd
replicated fig 2a

@@
 
-# acquire fig_1_data from gig_1.csv
+# acquire fig 1 data from fig_1.csv
 fig_1_data <- 
   read.csv(file=here("data/fig_1.csv"))
+
+# acquire fig 2 a data from figs_2a_2b.csv
+fig_2a_data <-
+  read.csv((file=here("data/figs_2a_2b.csv")))
+
+fig_2a_data <- fig_2a_data %>% select(year, brate_1519, brate_2024, brate_2529, brate_3034, brate_3539, brate_4044)
+
+fig_2a_data_long <- fig_2a_data %>%
+  pivot_longer(cols = -year, names_to = "Age_Group", values_to = "Birth_Rate")
 ```
@@
 
+```{r graph2, fig.cap="Trends in Birth Rates by Population Subgroup", echo=FALSE}
+
+#| label: fig-2
+
+# replicate figure 2a from the original paper
+fig_2a_data_long %>% ggplot(aes(x = year, y = Birth_Rate, color = Age_Group)) +
+  geom_line(size = 1) +
+  labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Age Group", caption = "Five-year age group") +
+  scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +
+  scale_x_continuous(limits = c(1980, 2020), breaks = seq(1980, 2020, by = 5)) +
+  geom_text(aes(x = 1985, y = 120), label = "Age 25-29", color = "black") +
+  geom_text(aes(x = 1985, y = 105), label = "Age 20-24", color = "black") +
+  geom_text(aes(x = 1985, y = 80), label = "Age 30-34", color = "black") +
+  geom_text(aes(x = 1985, y = 58), label = "Age 15-19", color = "black") +
+  geom_text(aes(x = 1985, y = 33), label = "Age 35-39", color = "black") +
+  geom_text(aes(x = 1985, y = 10), label = "Age 40-44", color = "black") +
+  theme_classic() +
+  theme(legend.position = "none")
+
+```
+
 # References

### 2024-02-12 paper/paper.qmd
replicated fig 2d

@@
   pivot_longer(cols = -year, names_to = "Age_Group", values_to = "Birth_Rate")
+
+# acquire fig 2 d data from figs_1_2_3.csv
+fig_2d_data <-
+  read.csv((file=here("data/figs_1_2_3.csv")))
+
+fig_2d_data <- fig_2d_data %>% select(year, brate_hsdropout,	brate_hsgrad,	brate_somecol,	brate_colgrad)
+
+fig_2d_data <- na.omit(fig_2d_data)
+
+fig_2d_data_long <- fig_2d_data %>%
+  pivot_longer(cols = -year, names_to = "Education_Level", values_to = "Birth_Rate")
 ```
@@
 
+# replicate figure 2d from the original paper
+fig_2d_data_long %>% ggplot(aes(x = year, y = Birth_Rate, color = Education_Level)) +
+  geom_line(size = 1) +
+  labs(x = "Year", y = "", color = "Education Level", caption = "Mother’s level of education (ages 20–44)") +
+  scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +
+  scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
+  geom_text(aes(x = 1995, y = 105), label = "No high school degree", color = "black") +
+  geom_text(aes(x = 1995, y = 75), label = "College graduate
+", color = "#d44215") +
+  theme_classic() +
+  geom_text(aes(x = 1995, y = 65), label = "High school graduate", color = "#0cc4be") +
+  geom_text(aes(x = 1995, y = 45), label = "Some college", color = "black") +
+  theme_classic() +
+  theme(legend.position = "none")
 ```

### 2024-02-12 paper/paper.qmd
slight readme update, added llm usage thus far, read fig 5 data

@@
 library(here)
+library(haven)
 
@@
   pivot_longer(cols = -year, names_to = "Education_Level", values_to = "Birth_Rate")
+
+# acquire fig 5 data from nchs_cohort_analysis.dta
+nchs_cohort_data <- 
+  read_dta(here("data/nchs_cohort_analysis.dta"))
 ```
@@
 
+```{r graph3, fig.cap="Children Ever Born by Mother’s Age and Birth Cohort", echo=FALSE}
+
+#| label: fig-3
+
+```
 # References

### 2024-02-12 paper/paper.qmd
small fix

@@
   geom_line(size = 1) +
-  labs(x = "Year", y = "", color = "Education Level", caption = "Mother’s level of education (ages 20–44)") +
+  labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Education Level", caption = "Mother’s level of education (ages 20–44)") +
   scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +

### 2024-02-12 scripts/99-replications.R
updated 99-replications.r

@@ -9,11 +9,82 @@
 
 
 #### Workspace setup ####
+
 library(tidyverse)
-# [...UPDATE THIS...]
+library(dplyr)
+library(ggplot2)
+library(here)
+library(haven)
 
 #### Load data ####
-# [...ADD CODE HERE...]
+# acquire fig 1 data from fig_1.csv
+fig_1_data <- 
+  read.csv(file=here("data/fig_1.csv"))
+
+# acquire fig 2 a data from figs_2a_2b.csv
+fig_2a_data <-
+  read.csv((file=here("data/figs_2a_2b.csv")))
+
+fig_2a_data <- fig_2a_data %>% select(year, brate_1519, brate_2024, brate_2529, brate_3034, brate_3539, brate_4044)
+
+fig_2a_data_long <- fig_2a_data %>%
+  pivot_longer(cols = -year, names_to = "Age_Group", values_to = "Birth_Rate")
+
+# acquire fig 2 d data from figs_1_2_3.csv
+fig_2d_data <-
+  read.csv((file=here("data/figs_1_2_3.csv")))
+
+fig_2d_data <- fig_2d_data %>% select(year, brate_hsdropout,	brate_hsgrad,	brate_somecol,	brate_colgrad)
+
+fig_2d_data <- na.omit(fig_2d_data)
+
+fig_2d_data_long <- fig_2d_data %>%
+  pivot_longer(cols = -year, names_to = "Education_Level", values_to = "Birth_Rate")
+
+# acquire fig 5 data from nchs_cohort_analysis.dta
+nchs_cohort_data <- 
+  read_dta(here("data/nchs_cohort_analysis.dta"))
+
+# replicate figure 1 from the original paper
+fig_1_data %>% ggplot(aes(x = year, y = brate_all)) +
+  geom_line(color = "#378bd4", size = 1) +
+  labs(x = "Year", y = "Births per 1,000 women age 15–44") +
+  scale_y_continuous(limits = c(50,80), breaks = seq(50, 80, by = 5)) +
+  scale_x_continuous(limits = c(1980,2020), breaks = seq(1980, 2020, by = 5)) +
+  theme_classic()
+
+# replicate figure 2a from the original paper
+fig_2a_data_long %>% ggplot(aes(x = year, y = Birth_Rate, color = Age_Group)) +
+  geom_line(size = 1) +
+  labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Age Group", caption = "Five-year age group") +
[…]

### 2024-02-12 paper/paper.qmd
updated qmd/pdf

@@
 format: pdf
+toc: true
 number-sections: true
@@
 ```{r graph1, fig.cap="Trend in US Birth Rates from 1980 - 2020", echo=FALSE}
+#| echo: false
+#| message: false
+#| warning: false
 
@@
 ```{r graph2, fig.cap="Trends in Birth Rates by Population Subgroup", echo=FALSE}
-
+#| echo: false
+#| message: false
+#| warning: false
 #| label: fig-2
@@
 ```{r graph3, fig.cap="Children Ever Born by Mother’s Age and Birth Cohort", echo=FALSE}
-
+#| echo: false
+#| message: false
+#| warning: false
 #| label: fig-3

### 2024-02-12 scripts/02-data_cleaning.R
added figure 2f replication

@@ -2,7 +2,7 @@
 # Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
 # Author: Aaron Xiaozhou Liu, Jimmy Luc, Maroosh Gillani
 # Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca, maroosh.gillani@mail.utoronto.ca
+# Contact: maroosh.gillani@mail.utoronto.ca, aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: [...UPDATE THIS...]
 # Any other information needed? [...UPDATE THIS...]

### 2024-02-12 scripts/03-test_data.R

@@ -2,7 +2,7 @@
 # Purpose: Tests... [...UPDATE THIS...]
 # Author: Aaron Xiaozhou Liu, Jimmy Luc, Maroosh Gillani
 # Date: 9 February 2024
-# Contact: aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca, maroosh.gillani@mail.utoronto.ca
+# Contact: maroosh.gillani@mail.utoronto.ca, aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: [...UPDATE THIS...]
 # Any other information needed? [...UPDATE THIS...]

### 2024-02-12 scripts/99-replications.R

@@ -1,11 +1,10 @@
 #### Preamble ####
-# Purpose: Replicated graphs from... [...UPDATE THIS...]
+# Purpose: Replication of figures in the original paper
 # Author: Aaron Xiaozhou Liu, Jimmy Luc, Maroosh Gillani
-# Date: 9 February 2024
-# Contact: aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca, maroosh.gillani@mail.utoronto.ca
+# Date: 12 February 2024
+# Contact: maroosh.gillani@mail.utoronto.ca, aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Must have data downloaded
 
 
 #### Workspace setup ####
@@ -17,43 +16,32 @@ library(here)
 library(haven)
 
 #### Load data ####
-# acquire fig 1 data from fig_1.csv
+
+### Figure 1 ###
+# acquire figure 1 data from fig_1.csv
 fig_1_data <- 
   read.csv(file=here("data/fig_1.csv"))
 
-# acquire fig 2 a data from figs_2a_2b.csv
+# replicate figure 1 from the original paper
+fig_1_data %>% ggplot(aes(x = year, y = brate_all)) +
+  geom_line(color = "#378bd4", size = 1) +
+  labs(x = "Year", y = "Births per 1,000 women age 15-44") +
+  scale_y_continuous(limits = c(50,80), breaks = seq(50, 80, by = 5)) +
+  scale_x_continuous(limits = c(1980,2020), breaks = seq(1980, 2020, by = 5)) +
+  theme_classic()
+
+### Figure 2 ###
+## Figure 2A ##
+# acquire figure 2A data from figs_2a_2b.csv
 fig_2a_data <-
-  read.csv((file=here("data/figs_2a_2b.csv")))
+  read.csv(file=here("data/figs_2a_2b.csv"))
 
 fig_2a_data <- fig_2a_data %>% select(year, brate_1519, brate_2024, brate_2529, brate_3034, brate_3539, brate_4044)
 
 fig_2a_data_long <- fig_2a_data %>%
   pivot_longer(cols = -year, names_to = "Age_Group", values_to = "Birth_Rate")
 
[…]

### 2024-02-12 paper/paper.qmd
Updated Title and abstract

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
+title: "A Deeper Analysis into the Trends and Factors of the US's Declining Birth Rates"
+#subtitle: "My subtitle if needed"
 author: 
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "Over the last couple of years, the American population has been witnessing a steady decline in their birth rates, as suggested by [@citeOriginalReplication]'s data from the American Economic Association.
+Through the analysis of multiple different factors, such as age and education, we notice a continual decline in birth rate starting from the year 2007.
+We explore how different social, cultural, and economic effects play a part in this worrying trend.
+This report sheds light on how the societal trends in the recent years have affected the American citizens in the last few decades, and how these patterns have played a part in this downward trend as people adapt to them."
 format: pdf

### 2024-02-13 paper/paper.qmd
Added Introduction

@@
 
+\newpage
 # Introduction
 
-# CHANGE THIS
+The United States has been experiencing heavy fluctuations in birth rates for nearly half a century with constant decline since 2007.
+The birth rate of a country is representative of its socioeconomic state, political state, and future prosperity.
+Furthermore, it determines the decision-making and distribution of government resources, education, health care services, and urban planning.
+This alarming issue leaves many seeking an explanation as to what exactly is causing this phenomenon.
+While people may point fingers at the ever changing societal norm, evolving ideology on family structure, or to no fault but the government, this paper aims to expose the possible explanations to the plummeting birth rate in the United States of America.
+
+The nations birth rate has fallen significantly since 2007, dropping but about 23% by 2022, measured by number of births per 1000 women [@citeAxios] and that birth rates tend to fall as based on the economy.
+2007 marks the start of the most severe economic crisis the world has ever faced, spanning the years 2007 until 2009.
+The events of this crisis has had long lasting damage on the economy, with unemployment doubling and the crash of the housing market [@citeFDR].
+These poor economic conditions and the unusually slow recovery of the economy would play a large role in the decision to have children, as an unpredictable economic environment makes people feel less secure about raising a family, thus contributing to the recorded decline in birth rates.
+This is just one of various factors that we will be exploring in this paper to observe any trends and correlations between societal and economic shifts and the dwindling birth rate in the US.
+
+Throughout the rest of this paper, we will be delving into the statistics with regards to birth rates in the US and observe any trends we can find.
+In the data and results sections of this paper, we examine and analyze the information reported by the AEA by downloading and cleaning the data sets.
+After that, we will look into the results sections to look for trends and other insights that can be uncovered from our research, and finally discuss and summarize all of our findings in the conclusion.
+
+# Data
+
+## Data source
 

### 2024-02-13 paper/paper.qmd
added figure 2f to quarto document

@@
 
-# acquire fig 2 a data from figs_2a_2b.csv
+# acquire fig 2a data from figs_2a_2b.csv
 fig_2a_data <-
@@
 
-# acquire fig 2 d data from figs_1_2_3.csv
+# acquire fig 2d data from figs_1_2_3.csv
 fig_2d_data <-
@@
   pivot_longer(cols = -year, names_to = "Education_Level", values_to = "Birth_Rate")
+
+# acquire figure 2F data from NCHS Natality Microdata's nchs_births_pop_1990_2019.dta and SEER's age_race_comp_seer.dta
+nchs_births_data <- read_dta(here("data/nchs_births_pop_1990_2019.dta"))
+seer_data <- read_dta(here("data/age_race_comp_seer.dta"))
+
+merged_data <- merge(nchs_births_data, seer_data, by = c("stname", "year"))
+
+merged_data <- merged_data %>%
+  mutate(
+    brate_firstbirth = merged_data$numbirth_firstbirth / merged_data$pop1544 * 1000,
+    brate_secondbirth = merged_data$numbirth_secondbirth / merged_data$pop1544 * 1000,
+    brate_thirdbirth = merged_data$numbirth_thirdbirth / merged_data$pop1544 * 1000,
+    brate_fourthbirth = merged_data$numbirth_fourthbirth / merged_data$pop1544 * 1000
+  ) %>%
+  select(year, starts_with("brate_"))
+
+merged_data_long <- merged_data %>%
+  pivot_longer(cols = -year, names_to = "Parity", values_to = "Birth_Rate")
 
@@
   theme(legend.position = "none")
+
+# replicate figure 2f from the original paper
+merged_data_long %>% ggplot(aes(x = year, y = round(Birth_Rate, digits = 1), color = Parity)) +
+  geom_line(size = 1, stat = "summary", fun = mean) +
+  labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Parity", caption = "Parity (ages 15-44)") +
+  scale_y_continuous(limits = c(0, 40), breaks = seq(0, 40, by = 10)) +
+  scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
+  geom_text(aes(x = 1995, y = 28), label = "First birth", color = "black") +
+  geom_text(aes(x = 1995, y = 23), label = "Second birth", color = "black") +
+  geom_text(aes(x = 1995, y = 12.5), label = "Third Birth", color = "black") +
+  geom_text(aes(x = 1995, y = 8.5), label = "Fourth+ birth", color = "black") +
+  theme_classic() +
+  theme(legend.position = "none")
 ```

### 2024-02-13 scripts/99-replications.R

@@ -88,7 +88,7 @@ fig_2d_data_long %>% ggplot(aes(x = year, y = Birth_Rate, color = Education_Leve
 nchs_births_data <- read_dta(here("data/nchs_births_pop_1990_2019.dta"))
 seer_data <- read_dta(here("data/age_race_comp_seer.dta"))
 
-merged_data <- merge(nchs_data, seer_data, by = c("stname", "year"))
+merged_data <- merge(nchs_births_data, seer_data, by = c("stname", "year"))
 
 merged_data <- merged_data %>%
   mutate(
@@ -108,10 +108,10 @@ merged_data_long %>% ggplot(aes(x = year, y = round(Birth_Rate, digits = 1), col
   labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Parity", caption = "Parity (ages 15-44)") +
   scale_y_continuous(limits = c(0, 40), breaks = seq(0, 40, by = 10)) +
   scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
-  geom_text(aes(x = 1995, y = 30), label = "First birth", color = "black") +
-  geom_text(aes(x = 1995, y = 25), label = "Second birth", color = "black") +
-  geom_text(aes(x = 1995, y = 15), label = "Third Birth", color = "black") +
-  geom_text(aes(x = 1995, y = 10), label = "Fourth+ birth", color = "black") +
+  geom_text(aes(x = 1995, y = 28), label = "First birth", color = "black") +
+  geom_text(aes(x = 1995, y = 23), label = "Second birth", color = "black") +
+  geom_text(aes(x = 1995, y = 12.5), label = "Third Birth", color = "black") +
+  geom_text(aes(x = 1995, y = 8.5), label = "Fourth+ birth", color = "black") +
   theme_classic() +
   theme(legend.position = "none")
 

### 2024-02-13 paper/paper.qmd
added data source information to document

@@
 
+The datasets for this paper were acquired through the American Economic Association's paper "The Puzzle of Falling US Birth Rates since the Great Recession" [@citeOriginalReplication], with the aid of data available from Centers for Disease Control and Prevention (CDC) Natality data [@citeCDCNCHS] and CDC SEER's U.S.
+Population Data [@citeCDCSeer].
+
 ```{r graph1, fig.cap="Trend in US Birth Rates from 1980 - 2020", echo=FALSE}

### 2024-02-13 paper/paper.qmd
added to data section of document

--- 2024-02-13 paper/paper.qmd
@@
 
-## Data source
+## Data Source and Collection
+
+The datasets for this paper were acquired through the American Economic Association's paper "The Puzzle of Falling US Birth Rates since the Great Recession" [@citeOriginalReplication], with the aid of data available from Centers for Disease Control and Prevention (CDC) Natality data [@citeCDCNCHS] and CDC SEER U.S.
+Population Data [@citeCDCSeer].
+The data sources were accessed on February 12th, 2024 and acts as the basis of this paper.
+The data sources themselves were a collaborative effort from the U.S.
+Census Bureau, the National Cancer Institute, SEER Data, CDC, and the National Center for Disease Control and Prevention (NCHS).
+The U.S.
+census is legally mandated by the United States Constitution which occurs every 10 years.
+The data is largely comprised of `numeric` values representing years, birth rates by race and ethnicity, marital status, education, age, and birth parity, or `character` values representing US states.
+
+All data collection and analysis was done using statistical tools and data visualization program R [@citeR] and Rstudio IDE [@citeRStudio] to help streamline the workflow.
+The analysis was done using the R program and the following supporting packages `tidyverse` [@citeTidyverse], `haven` [@citeHaven], `ggplot2` [@citeGgplot2], `here` [@citeHere], `ggthemes` [@citeGgthemes], and `dplyr` [@citeDplyr].
+We will provide more comprehensive details about the data cleaning, and analysis in the following subsections.
+
+## Variables of Interest
+Although the data contains an abundance of information, we are mainly concerned of US birth rates per year, the total female population, and various other factors including age, mother's level of education, and birth parity (number of children for a given women).
+As this paper focuses on birth rates in the US over the course of time, we have all the necessary and sufficient information required to proceed processing the data for our analysis.
+
+## Data Processing
+
+## Data Visualization
 
@@
   geom_line(color = "#378bd4", size = 1) +
-  labs(x = "Year", y = "Births per 1,000 women age 15–44") +
+  labs(x = "Year", y = "Births per 1,000 women age 15-44") +
   scale_y_continuous(limits = c(50,80), breaks = seq(50, 80, by = 5)) +
@@
   geom_line(size = 1) +
-  labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Education Level", caption = "Mother’s level of education (ages 20–44)") +
+  labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Education Level", caption = "Mother's level of education (ages 20-44)") +
   scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +
@@
 
-```{r graph3, fig.cap="Children Ever Born by Mother’s Age and Birth Cohort", echo=FALSE}
+```{r graph3, fig.cap="Children Ever Born by Mother's Age and Birth Cohort", echo=FALSE}
 #| echo: false

### 2024-02-13 paper/paper.qmd
update minor tweaks

@@
 
-The datasets for this paper were acquired through the American Economic Association's paper "The Puzzle of Falling US Birth Rates since the Great Recession" [@citeOriginalReplication], with the aid of data available from Centers for Disease Control and Prevention (CDC) Natality data [@citeCDCNCHS] and CDC SEER U.S.
+The datasets for this paper were acquired through the American Economic Association's paper "The Puzzle of Falling US Birth Rates since the Great Recession" [@citeOriginalReplication], with the aid of data available from Centers for Disease Control, and Prevention (CDC) Natality data [@citeCDCNCHS] and CDC SEER U.S.
 Population Data [@citeCDCSeer].
-The data sources were accessed on February 12th, 2024 and acts as the basis of this paper.
-The data sources themselves were a collaborative effort from the U.S.
-Census Bureau, the National Cancer Institute, SEER Data, CDC, and the National Center for Disease Control and Prevention (NCHS).
-The U.S.
-census is legally mandated by the United States Constitution which occurs every 10 years.
+The data sources were accessed on February 12th, 2024 and acts as the foundation of this paper.
+The data sources themselves are a collaborative effort including but not limited to the U.S.
+Census Bureau, the National Cancer Institute, SEER Data, CDC, and the National Center for Health Statistics (NCHS).
+The United States census is legally mandated by the United States Constitution which occurs every 10 years.
 The data is largely comprised of `numeric` values representing years, birth rates by race and ethnicity, marital status, education, age, and birth parity, or `character` values representing US states.

### 2024-02-13 paper/paper.qmd
update figure text positioning

@@
 This report sheds light on how the societal trends in the recent years have affected the American citizens in the last few decades, and how these patterns have played a part in this downward trend as people adapt to them."
-format: pdf
+format:
+  pdf:
+    fig-pos: "H"
 toc: true
@@
   geom_text(aes(x = 1985, y = 120), label = "Age 25-29", color = "black") +
-  geom_text(aes(x = 1985, y = 105), label = "Age 20-24", color = "black") +
+  geom_text(aes(x = 1985, y = 100), label = "Age 20-24", color = "black") +
   geom_text(aes(x = 1985, y = 80), label = "Age 30-34", color = "black") +
@@
   scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
-  geom_text(aes(x = 1995, y = 105), label = "No high school degree", color = "black") +
-  geom_text(aes(x = 1995, y = 75), label = "College graduate
-", color = "#d44215") +
+  geom_text(aes(x = 1993, y = 110), label = "No high school degree", color = "black", hjust = 0) +
+  geom_text(aes(x = 1993, y = 82), label = "College graduate", color = "#f8766d", hjust = 0) +
   theme_classic() +
-  geom_text(aes(x = 1995, y = 65), label = "High school graduate", color = "#0cc4be") +
-  geom_text(aes(x = 1995, y = 45), label = "Some college", color = "black") +
+  geom_text(aes(x = 1993, y = 62), label = "High school graduate", color = "#00bfc4", hjust = 0) +
+  geom_text(aes(x = 1993, y = 45), label = "Some college", color = "black", hjust = 0) +
   theme_classic() +
@@
   scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
-  geom_text(aes(x = 1995, y = 28), label = "First birth", color = "black") +
-  geom_text(aes(x = 1995, y = 23), label = "Second birth", color = "black") +
-  geom_text(aes(x = 1995, y = 12.5), label = "Third Birth", color = "black") +
-  geom_text(aes(x = 1995, y = 8.5), label = "Fourth+ birth", color = "black") +
+  geom_text(aes(x = 1993, y = 28), label = "First birth", color = "black", hjust = 0) +
+  geom_text(aes(x = 1993, y = 23), label = "Second birth", color = "black", hjust = 0) +
+  geom_text(aes(x = 1993, y = 12.5), label = "Third Birth", color = "black", hjust = 0) +
+  geom_text(aes(x = 1993, y = 8.5), label = "Fourth+ birth", color = "black", hjust = 0) +
   theme_classic() +

### 2024-02-13 paper/paper.qmd
update figures in document

@@
 
-```{r graph2, fig.cap="Trends in Birth Rates by Population Subgroup", echo=FALSE}
+```{r graph2a, fig.cap="Trends in Birth Rates by Five-year Age Group", echo=FALSE}
 #| echo: false
@@
   geom_line(size = 1) +
-  labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Age Group", caption = "Five-year age group") +
+  labs(x = "Year", y = "Births per 1,000 women in \n relevant population subgroup", color = "Age Group") +
   scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +
@@
   theme(legend.position = "none")
+```
 
+```{r graph2b, fig.cap="Trends in Birth Rates by Mother's Level of Education (ages 20-44)", echo=FALSE}
 # replicate figure 2d from the original paper
@@
   geom_line(size = 1) +
-  labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Education Level", caption = "Mother's level of education (ages 20-44)") +
+  labs(x = "Year", y = "Births per 1,000 women in \n relevant population subgroup", color = "Education Level") +
   scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +
   scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
-  geom_text(aes(x = 1993, y = 110), label = "No high school degree", color = "black", hjust = 0) +
+  geom_text(aes(x = 1993, y = 112), label = "No high school degree", color = "black", hjust = 0) +
   geom_text(aes(x = 1993, y = 82), label = "College graduate", color = "#f8766d", hjust = 0) +
@@
   theme(legend.position = "none")
+```
+```{r graph2c, fig.cap="Trends in Birth Rates by Parity (ages 15-44)", echo=FALSE}
 
@@
   geom_line(size = 1, stat = "summary", fun = mean) +
-  labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Parity", caption = "Parity (ages 15-44)") +
+  labs(x = "Year", y = "Births per 1,000 women in \n relevant population subgroup", color = "Parity") +
   scale_y_continuous(limits = c(0, 40), breaks = seq(0, 40, by = 10)) +

### 2024-02-13 paper/paper.qmd
added figure 5

@@
 ---
-```{r, include=FALSE, warning=FALSE, message=FALSE, echo=FALSE}
+```{r, include=FALSE, warning=FALSE, message=FALSE}
 # workplace setup
@@
 # acquire fig 5 data from nchs_cohort_analysis.dta
+# acquire figure 5 data from CDC SEER's agecomp-seer.dta
+seer_cohort_data <- 
+  read_dta(here("data/agecomp-seer.dta"))
+
 nchs_cohort_data <- 
   read_dta(here("data/nchs_cohort_analysis.dta"))
+
+# code to replicate figure 5 from: https://github.com/KCtt457/us-birth-rates-decline
+# Process population data to aggregate by cohort
+fig5_pop_data1 <- seer_cohort_data %>%
+  select(1:32) %>%
+  select(-c(stname)) %>%
+  group_by(year) %>%
+  summarise(across(everything(),sum)) # https://stackoverflow.com/questions/1660124/how-to-sum-a-variable-by-group
+
+fig5_pop_data1 <- fig5_pop_data1 %>%
+  pivot_longer(-c(year), names_to = "mage", values_to = "pop")
+
+fig5_pop_data1 <- fig5_pop_data1 %>%
+  mutate(mage = as.numeric(str_replace(mage, "fem", ""))) %>%
+  mutate(cohort = year - mage) %>%
+  mutate(cohort2 = case_when(cohort >= 1968 & cohort <=1972 ~ 1,
+                             cohort >= 1973 & cohort <=1977 ~ 2,
+                             cohort >= 1978 & cohort <=1982 ~ 3,
+                             cohort >= 1983 & cohort <=1987 ~ 4,
+                             cohort >= 1988 & cohort <=1992 ~ 5,
+                             cohort >= 1993 & cohort <=1997 ~ 6)
+         )
+
+fig5_pop_data1 <- na.omit(fig5_pop_data1)
+fig5_pop_data1 <- fig5_pop_data1 %>%
+  group_by(cohort2, mage) %>%
+  summarise(pop = sum(pop))
+
+# Join dataframes for population and birth data
+nchs_cohort_data <- full_join(nchs_cohort_data, fig5_pop_data1, by=c("mage", "cohort2")) %>%
+  mutate(cohort_years = case_when(cohort2 == 1 ~ "1968-1972",
+                                  cohort2 == 2 ~ "1973-1977",
+                                  cohort2 == 3 ~ "1978-1982",
+                                  cohort2 == 4 ~ "1983-1987",
+                                  cohort2 == 5 ~ "1988-1992",
+                                  cohort2 == 6 ~ "1993-1997")
+         )
+
+# Calculate birth rates
+nchs_cohort_data <- nchs_cohort_data %>%
+  mutate(brate=numbirth/pop*1000,
+         age_20_24_year = case_when(cohort2 == 1 ~ 1992,
+                                    cohort2 == 2 ~ 1997,
+                                    cohort2 == 3 ~ 2002,
+                                    cohort2 == 4 ~ 2007,
+                                    cohort2 == 5 ~ 2012,
+                                    cohort2 == 6 ~ 2017)
+         )
+
+# Calculate cumulative birth rates for each cohort
+nchs_cohort_data <- nchs_cohort_data %>%
+  group_by(cohort2) %>%
+  mutate(cum_brate = cumsum(brate)/1000, 
+         cohort_years = case_when(cohort2 == 1 ~ "1968-1972",
+                                  cohort2 == 2 ~ "1973-1977",
+                                  cohort2 == 3 ~ "1978-1982",
+                                  cohort2 == 4 ~ "1983-1987",
+                                  cohort2 == 5 ~ "1988-1992",
+                                  cohort2 == 6 ~ "1993-1997")
+         )
+
 ```
@@
 
-```{r graph1, fig.cap="Trend in US Birth Rates from 1980 - 2020", echo=FALSE}
+```{r graph1, fig.cap="Trend in US Birth Rates from 1980 - 2020"}
 #| echo: false
@@
 #| warning: false
-
 #| label: fig-1
@@
 
-```{r graph2a, fig.cap="Trends in Birth Rates by Five-year Age Group", echo=FALSE}
+```{r graph2a, fig.cap="Trends in Birth Rates by Five-year Age Group"}
 #| echo: false
@@
 #| warning: false
-#| label: fig-2
+#| label: fig-2a
 
@@
 
-```{r graph2b, fig.cap="Trends in Birth Rates by Mother's Level of Education (ages 20-44)", echo=FALSE}
+```{r graph2b, fig.cap="Trends in Birth Rates by Mother's Level of Education (ages 20-44)"}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-2b
 # replicate figure 2d from the original paper
@@
 ```
-```{r graph2c, fig.cap="Trends in Birth Rates by Parity (ages 15-44)", echo=FALSE}
+
+```{r graph2c, fig.cap="Trends in Birth Rates by Parity (ages 15-44)"}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-2c
 
@@
 
-```{r graph3, fig.cap="Children Ever Born by Mother's Age and Birth Cohort", echo=FALSE}
+```{r graph3, fig.cap="Children Ever Born by Mother's Age and Birth Cohort"}
 #| echo: false
@@
 
+# replicate figure 5 from the original paper
+# code to replicate figure 5 from: https://github.com/KCtt457/us-birth-rates-decline
+nchs_cohort_data %>%
+  ggplot(aes(x = mage,
+             y = cum_brate,
+             color = cohort_years)) +
+  geom_line(size = 1) +
+  labs(x = "Mother's age", y = "Children ever born", color = "Birth cohorts") +
+  scale_y_continuous(limits=c(0, 2.5), breaks=seq(0, 2.5, by = .5)) +
+  scale_x_continuous(limits=c(15, 44), breaks=seq(15, 44, by = 1)) +
+  theme_classic() + 
+  theme(legend.background = element_blank(), legend.box.background = element_rect(), legend.position = c(0.85,0.4), axis.text.x = element_text(angle = 45, hjust = 1))
 ```

### 2024-02-13 scripts/99-replications.R

@@ -31,8 +31,8 @@ fig_1_data %>% ggplot(aes(x = year, y = brate_all)) +
   theme_classic()
 
 ### Figure 2 ###
-## Figure 2A ##
-# acquire figure 2A data from figs_2a_2b.csv
+## Figure 2a ##
+# acquire figure 2a data from figs_2a_2b.csv
 fig_2a_data <-
   read.csv(file=here("data/figs_2a_2b.csv"))
 
@@ -41,14 +41,14 @@ fig_2a_data <- fig_2a_data %>% select(year, brate_1519, brate_2024, brate_2529, 
 fig_2a_data_long <- fig_2a_data %>%
   pivot_longer(cols = -year, names_to = "Age_Group", values_to = "Birth_Rate")
 
-# replicate figure 2A from the original paper
+# replicate figure 2a from the original paper
 fig_2a_data_long %>% ggplot(aes(x = year, y = Birth_Rate, color = Age_Group)) +
   geom_line(size = 1) +
-  labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Age Group", caption = "Five-year age group") +
+  labs(x = "Year", y = "Births per 1,000 women in \n relevant population subgroup", color = "Age Group") +
   scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +
   scale_x_continuous(limits = c(1980, 2020), breaks = seq(1980, 2020, by = 5)) +
   geom_text(aes(x = 1985, y = 120), label = "Age 25-29", color = "black") +
-  geom_text(aes(x = 1985, y = 105), label = "Age 20-24", color = "black") +
+  geom_text(aes(x = 1985, y = 100), label = "Age 20-24", color = "black") +
   geom_text(aes(x = 1985, y = 80), label = "Age 30-34", color = "black") +
   geom_text(aes(x = 1985, y = 58), label = "Age 15-19", color = "black") +
   geom_text(aes(x = 1985, y = 33), label = "Age 35-39", color = "black") +
@@ -56,8 +56,8 @@ fig_2a_data_long %>% ggplot(aes(x = year, y = Birth_Rate, color = Age_Group)) +
   theme_classic() +
   theme(legend.position = "none")
 
-## Figure 2D ##
-# acquire figure 2D data from figs_1_2_3.csv
+## Figure 2d ##
+# acquire figure 2d data from figs_1_2_3.csv
[…]

### 2024-02-13 paper/paper.qmd
updated paper.qmd and pdf

@@
 #subtitle: "My subtitle if needed"
-author: 
-  - Maroosh Gillani
-  - Aaron Liu
-  - Jimmy Luc
-thanks: "Code and data are available at: https://github.com/rex009x/us_birth_rates/tree/master/data . A replication of various aspects in this paper are available at: LINK ."
+author: Maroosh Gillani, Aaron Liu & Jimmy Luc
+thanks: "Code and data are available at: https://github.com/rex009x/us_birth_rates/tree/master/data.
+Selected aspects of the original paper that underpins this paper were reproduced and a record of this is available at: https://www.socialsciencereproduction.org/reproductions/bfe97499-4c05-466f-8cb5-fbedc27a83a1/index "
 date: today

### 2024-02-13 scripts/03-test_data.R
added tests for data types

@@ -1,15 +1,65 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
+# Purpose: Tests data acquired from original AEA paper's replication package
 # Author: Aaron Xiaozhou Liu, Jimmy Luc, Maroosh Gillani
-# Date: 9 February 2024
+# Date: 13 February 2024
 # Contact: maroosh.gillani@mail.utoronto.ca, aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Must have data downloaded
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(here)
+library(haven)
+
+#### Load in data ####
+fig_1_data = read.csv(file=here("data/fig_1.csv"))
+fig_2a_data = read.csv(file=here("data/figs_2a_2b.csv"))
+fig_2d_data = read.csv(file=here("data/figs_1_2_3.csv"))
+fig_2f_data_1 = read_dta(here("data/nchs_births_pop_1990_2019.dta"))
+fig_2f_data_2 = read_dta(here("data/age_race_comp_seer.dta"))
+fig_5_data_1 = read_dta(here("data/agecomp-seer.dta"))
+fig_5_data_2 = read_dta(here("data/nchs_cohort_analysis.dta"))
 
 #### Test data ####
+## test data types
+# methods used: https://stackoverflow.com/a/29591221
+all(sapply(fig_1_data$year, is.numeric))
+all(sapply(fig_1_data$brate_all, is.numeric))
+
+all(sapply(fig_2a_data$year, is.numeric))
+all(sapply(fig_2a_data$brate_1519, is.numeric))
+all(sapply(fig_2a_data$brate_2024, is.numeric))
+all(sapply(fig_2a_data$brate_2529, is.numeric))
+all(sapply(fig_2a_data$brate_3034, is.numeric))
+all(sapply(fig_2a_data$brate_3539, is.numeric))
+all(sapply(fig_2a_data$brate_4044, is.numeric))
+
+all(sapply(fig_2d_data$year, is.numeric))
+all(sapply(fig_2d_data$brate_hsdropout, is.numeric)) # is.numeric ignores NA values
[…]

### 2024-02-13 scripts/04-model.R

@@ -1,35 +1,0 @@
-#### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Aaron Xiaozhou Liu, Jimmy Luc, Maroosh Gillani
-# Date: 9 February 2024
-# Contact: maroosh.gillani@mail.utoronto.ca, aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(rstanarm)
-
-#### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
-
-### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
-
-
-#### Save model ####
-saveRDS(
-  first_model,
-  file = "models/first_model.rds"
-)

### 2024-02-13 scripts/03-test_data.R
added tests for data values

@@ -22,7 +22,7 @@ fig_5_data_1 = read_dta(here("data/agecomp-seer.dta"))
 fig_5_data_2 = read_dta(here("data/nchs_cohort_analysis.dta"))
 
 #### Test data ####
-## test data types
+## test data types##
 # methods used: https://stackoverflow.com/a/29591221
 all(sapply(fig_1_data$year, is.numeric))
 all(sapply(fig_1_data$brate_all, is.numeric))
@@ -61,5 +61,43 @@ all(sapply(fig_5_data_2$cohort2, is.numeric))
 all(sapply(fig_5_data_2$numbirth, is.numeric))
 
 
-## test data values
-min(fig_1_data$year) == 1980 && max(fig_1_data$year) == 2020
+## test data values ##
+all(fig_1_data$year >= 1980 & fig_1_data$year <= 2020)
+all(fig_1_data$brate_all >= 0) # cannot have negative birth rate
+
+
+all(fig_2a_data$year >= 1980 & fig_2a_data$year <= 2020)
+all(fig_2a_data$brate_1519 >= 0) # cannot have negative birth rate
+all(fig_2a_data$brate_2024 >= 0)
+all(fig_2a_data$brate_2529 >= 0)
+all(fig_2a_data$brate_3034 >= 0)
+all(fig_2a_data$brate_3539 >= 0)
+all(fig_2a_data$brate_4044 >= 0)
+
+# data contains empty entries
+all(fig_2d_data$year >= 1980 & fig_2d_data$year <= 2020, na.rm = TRUE)
+all(fig_2d_data$brate_hsdropout >= 0, na.rm = TRUE) # cannot have negative birth rate
+all(fig_2d_data$brate_hsgrad >= 0, na.rm = TRUE)
+all(fig_2d_data$brate_somecol >= 0, na.rm = TRUE)
+all(fig_2d_data$brate_colgrad >= 0, na.rm = TRUE)
+
+
+all(fig_2f_data_1$year >= 1990 & fig_2f_data_1$year <= 2019)
+all(fig_2f_data_1$stname != '')
+all(fig_2f_data_1$numbirth_firstbirth >= 0) # cannot have negative number of births
+all(fig_2f_data_1$numbirth_secondbirth >= 0)
+all(fig_2f_data_1$numbirth_thirdbirth >= 0)
+all(fig_2f_data_1$numbirth_fourthbirth >= 0)
[…]

### 2024-02-13 scripts/00-simulate_data.R
added simulated data

@@ -1,16 +1,68 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
+# Purpose: Simulates the data required to reproduce graphs
 # Author: Aaron Xiaozhou Liu, Jimmy Luc, Maroosh Gillani
-# Date: 9 February 2024
+# Date: 13 February 2024
 # Contact: maroosh.gillani@mail.utoronto.ca, aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
+# set seed for simulation
+set.seed(416)
+
+# simulate figure 1 data
+# methods used: https://stackoverflow.com/a/68763213
+fig_1_years = 1980:2020
+fig_1_brate_mean = 65
+fig_1_brate_sd = 2
+simulate_fig_1 <-
+  tibble(
+    year = fig_1_years,
+    brate_all = rnorm(n = fig_1_years, mean = fig_1_brate_mean, sd = fig_1_brate_sd)
+    )
+
+# simulate figure 2a data
+fig_2a_years = 1980:2020
+simulate_fig_2a <-
+  tibble(
+    year = fig_2a_years,
+    brate_1519 = rnorm(fig_2a_years, mean = 40, sd = 7),
+    brate_2024 = rnorm(fig_2a_years, mean = 100, sd = 7),
+    brate_2529 = rnorm(fig_2a_years, mean = 110, sd = 5),
+    brate_3034 = rnorm(fig_2a_years, mean = 90, sd = 10),
+    brate_3539 = rnorm(fig_2a_years, mean = 40, sd = 10),
+    brate_4044 = rnorm(fig_2a_years, mean = 5, sd = 5)
+  )
+
+# simulate figure 2d data
+fig_2d_years = 1990:2020
+simulate_fig_2d <-
+  tibble(
+    year = fig_2d_years,
+    brate_hsdropout = rnorm(fig_2d_years, mean = 105, sd = 12),
+    brate_hsgrad = rnorm(fig_2d_years, mean = 75, sd = 5),
[…]

### 2024-02-13 scripts/01-download_data.R

@@ -1,24 +1,0 @@
-#### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Aaron Xiaozhou Liu, Jimmy Luc, Maroosh Gillani
-# Date: 9 February 2024
-# Contact: maroosh.gillani@mail.utoronto.ca, aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
-
-#### Workspace setup ####
-library(opendatatoronto)
-library(tidyverse)
-# [...UPDATE THIS...]
-
-#### Download data ####
-# [...ADD CODE HERE TO DOWNLOAD...]
-
-
-
-#### Save data ####
-# [...UPDATE THIS...]
-# change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(the_raw_data, "inputs/data/raw_data.csv")

### 2024-02-13 scripts/01-test_data.R

@@ -5,6 +5,7 @@
 # Contact: maroosh.gillani@mail.utoronto.ca, aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: Must have data downloaded
+# Additional Information: these tests are applicable to the simulated data in 00-simulate_data.R
 
 
 #### Workspace setup ####

### 2024-02-13 scripts/02-data_cleaning.R

@@ -1,44 +1,0 @@
-#### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Aaron Xiaozhou Liu, Jimmy Luc, Maroosh Gillani
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: maroosh.gillani@mail.utoronto.ca, aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
-#### Workspace setup ####
-library(tidyverse)
-
-#### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
-
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
-  mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
-  ) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "490",
-                                 "49",
-                                 wing_width_mm)) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "6",
-                                 "60",
-                                 wing_width_mm)) |>
-  mutate(
-    wing_width_mm = as.numeric(wing_width_mm),
-    wing_length_mm = as.numeric(wing_length_mm),
-    flying_time_sec_first_timer = as.numeric(flying_time_sec_first_timer)
-  ) |>
-  rename(flying_time = flying_time_sec_first_timer,
-         width = wing_width_mm,
-         length = wing_length_mm
-         ) |> 
-  tidyr::drop_na()
-
-#### Save data ####
-write_csv(cleaned_data, "outputs/data/analysis_data.csv")

### 2024-02-14 paper/paper.qmd
update colour labels

@@
   scale_x_continuous(limits = c(1980, 2020), breaks = seq(1980, 2020, by = 5)) +
-  geom_text(aes(x = 1985, y = 120), label = "Age 25-29", color = "black") +
-  geom_text(aes(x = 1985, y = 100), label = "Age 20-24", color = "black") +
-  geom_text(aes(x = 1985, y = 80), label = "Age 30-34", color = "black") +
-  geom_text(aes(x = 1985, y = 58), label = "Age 15-19", color = "black") +
-  geom_text(aes(x = 1985, y = 33), label = "Age 35-39", color = "black") +
-  geom_text(aes(x = 1985, y = 10), label = "Age 40-44", color = "black") +
+  geom_text(aes(x = 1985, y = 120), label = "Age 25-29", color = "#1e7837") +
+  geom_text(aes(x = 1985, y = 100), label = "Age 20-24", color = "#786d1e") +
+  geom_text(aes(x = 1985, y = 80), label = "Age 30-34", color = "#27a39b") +
+  geom_text(aes(x = 1985, y = 58), label = "Age 15-19", color = "#f8766d") +
+  geom_text(aes(x = 1985, y = 33), label = "Age 35-39", color = "#5587cf") +
+  geom_text(aes(x = 1985, y = 10), label = "Age 40-44", color = "#d45dce") +
   theme_classic() +
@@
   scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
-  geom_text(aes(x = 1993, y = 112), label = "No high school degree", color = "black", hjust = 0) +
+  geom_text(aes(x = 1993, y = 112), label = "No high school degree", color = "#668a38", hjust = 0) +
   geom_text(aes(x = 1993, y = 82), label = "College graduate", color = "#f8766d", hjust = 0) +
@@
   geom_text(aes(x = 1993, y = 62), label = "High school graduate", color = "#00bfc4", hjust = 0) +
-  geom_text(aes(x = 1993, y = 45), label = "Some college", color = "black", hjust = 0) +
+  geom_text(aes(x = 1993, y = 45), label = "Some college", color = "#b873c7", hjust = 0) +
   theme_classic() +
@@
   scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
-  geom_text(aes(x = 1993, y = 28), label = "First birth", color = "black", hjust = 0) +
-  geom_text(aes(x = 1993, y = 23), label = "Second birth", color = "black", hjust = 0) +
-  geom_text(aes(x = 1993, y = 12.5), label = "Third Birth", color = "black", hjust = 0) +
-  geom_text(aes(x = 1993, y = 8.5), label = "Fourth+ birth", color = "black", hjust = 0) +
+  geom_text(aes(x = 1993, y = 28), label = "First birth", color = "#f8766d", hjust = 0) +
+  geom_text(aes(x = 1993, y = 23), label = "Second birth", color = "#00bfc4", hjust = 0) +
+  geom_text(aes(x = 1993, y = 12.5), label = "Third Birth", color = "#b873c7", hjust = 0) +
+  geom_text(aes(x = 1993, y = 8.5), label = "Fourth+ birth", color = "#668a38", hjust = 0) +
   theme_classic() +

### 2024-02-14 scripts/99-replications.R
label colour

@@ -47,12 +47,12 @@ fig_2a_data_long %>% ggplot(aes(x = year, y = Birth_Rate, color = Age_Group)) +
   labs(x = "Year", y = "Births per 1,000 women in \n relevant population subgroup", color = "Age Group") +
   scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +
   scale_x_continuous(limits = c(1980, 2020), breaks = seq(1980, 2020, by = 5)) +
-  geom_text(aes(x = 1985, y = 120), label = "Age 25-29", color = "black") +
-  geom_text(aes(x = 1985, y = 100), label = "Age 20-24", color = "black") +
-  geom_text(aes(x = 1985, y = 80), label = "Age 30-34", color = "black") +
-  geom_text(aes(x = 1985, y = 58), label = "Age 15-19", color = "black") +
-  geom_text(aes(x = 1985, y = 33), label = "Age 35-39", color = "black") +
-  geom_text(aes(x = 1985, y = 10), label = "Age 40-44", color = "black") +
+  geom_text(aes(x = 1985, y = 120), label = "Age 25-29", color = "#1e7837") +
+  geom_text(aes(x = 1985, y = 100), label = "Age 20-24", color = "#786d1e") +
+  geom_text(aes(x = 1985, y = 80), label = "Age 30-34", color = "#27a39b") +
+  geom_text(aes(x = 1985, y = 58), label = "Age 15-19", color = "#f8766d") +
+  geom_text(aes(x = 1985, y = 33), label = "Age 35-39", color = "#5587cf") +
+  geom_text(aes(x = 1985, y = 10), label = "Age 40-44", color = "#d45dce") +
   theme_classic() +
   theme(legend.position = "none")
 
@@ -74,11 +74,11 @@ fig_2d_data_long %>% ggplot(aes(x = year, y = Birth_Rate, color = Education_Leve
   labs(x = "Year", y = "Births per 1,000 women in \n relevant population subgroup", color = "Education Level") +
   scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +
   scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
-  geom_text(aes(x = 1993, y = 112), label = "No high school degree", color = "black", hjust = 0) +
+  geom_text(aes(x = 1993, y = 112), label = "No high school degree", color = "#668a38", hjust = 0) +
[…]

### 2024-02-14 paper/paper.qmd
fix merge conflicts

--- 2024-02-13 paper/paper.qmd
@@
 ---
-title: "A Deeper Analysis into the Trends and Factors of the US's Declining Birth Rates"
+title: "From Trends to Truth: A Comprehensive Analysis Revealing Contributing Factors Behind Declining US Birth Rates"
 #subtitle: "My subtitle if needed"
-author: 
-  - Maroosh Gillani
-  - Aaron Liu
-  - Jimmy Luc
-thanks: "Code and data are available at: https://github.com/rex009x/us_birth_rates/tree/master/data . A replication of various aspects in this paper are available at: LINK ."
+author: Maroosh Gillani, Aaron Liu & Jimmy Luc
+thanks: "Code and data are available at: https://github.com/rex009x/us_birth_rates/tree/master/data.
+Selected aspects of the original paper that underpins this paper were reproduced and a record of this is available at: https://www.socialsciencereproduction.org/reproductions/bfe97499-4c05-466f-8cb5-fbedc27a83a1/index "
 date: today
 date-format: long
-abstract: "Over the last couple of years, the American population has been witnessing a steady decline in their birth rates, as suggested by [@citeOriginalReplication]'s data from the American Economic Association.
+abstract: "Over the last couple of years, the American population has been witnessing a steady decline in their birth rates, as suggested by the American Economic Association's article 'The Puzzle of Falling US Birth Rates since the Great Recession' by Kearney, Levine, and Pardue (2020).
 Through the analysis of multiple different factors, such as age and education, we notice a continual decline in birth rate starting from the year 2007.
-We explore how different social, cultural, and economic effects play a part in this worrying trend.
-This report sheds light on how the societal trends in the recent years have affected the American citizens in the last few decades, and how these patterns have played a part in this downward trend as people adapt to them."
+We discover how different social, cultural, and economic shifts play a part in this worrying trend.
+This paper demonstrates how socioeconomic trends in the recent years have affected Americans, and how these patterns represented through data have played a key role in the irreversible downward trajectory of US birth rates."
 format:
@@
 # workplace setup
-
 library(tidyverse)
@@
 Throughout the rest of this paper, we will be delving into the statistics with regards to birth rates in the US and observe any trends we can find.
-In the data and results sections of this paper, we examine and analyze the information reported by the AEA by downloading and cleaning the data sets.
+In the data and results sections of this paper, we examine, analyze, and reproduce the information reported by the authors by downloading and cleaning the data sets provided in the replication package.
 After that, we will look into the results sections to look for trends and other insights that can be uncovered from our research, and finally discuss and summarize all of our findings in the conclusion.
@@
   scale_x_continuous(limits = c(1980, 2020), breaks = seq(1980, 2020, by = 5)) +
-  geom_text(aes(x = 1985, y = 120), label = "Age 25-29", color = "black") +
-  geom_text(aes(x = 1985, y = 100), label = "Age 20-24", color = "black") +
-  geom_text(aes(x = 1985, y = 80), label = "Age 30-34", color = "black") +
-  geom_text(aes(x = 1985, y = 58), label = "Age 15-19", color = "black") +
-  geom_text(aes(x = 1985, y = 33), label = "Age 35-39", color = "black") +
-  geom_text(aes(x = 1985, y = 10), label = "Age 40-44", color = "black") +
+  geom_text(aes(x = 1985, y = 120), label = "Age 25-29", color = "#00ba38") +
+  geom_text(aes(x = 1985, y = 100), label = "Age 20-24", color = "#b79f00") +
+  geom_text(aes(x = 1985, y = 80), label = "Age 30-34", color = "#00bfc4") +
+  geom_text(aes(x = 1985, y = 58), label = "Age 15-19", color = "#f8766d") +
+  geom_text(aes(x = 1985, y = 33), label = "Age 35-39", color = "#619cff") +
+  geom_text(aes(x = 1985, y = 10), label = "Age 40-44", color = "#f564e3") +
   theme_classic() +
@@
 #| label: fig-2b
+
 # replicate figure 2d from the original paper
@@
   scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
-  geom_text(aes(x = 1993, y = 112), label = "No high school degree", color = "black", hjust = 0) +
+  geom_text(aes(x = 1993, y = 112), label = "No high school degree", color = "#7cae00", hjust = 0) +
   geom_text(aes(x = 1993, y = 82), label = "College graduate", color = "#f8766d", hjust = 0) +
@@
   geom_text(aes(x = 1993, y = 62), label = "High school graduate", color = "#00bfc4", hjust = 0) +
-  geom_text(aes(x = 1993, y = 45), label = "Some college", color = "black", hjust = 0) +
+  geom_text(aes(x = 1993, y = 45), label = "Some college", color = "#c77cff", hjust = 0) +
   theme_classic() +
@@
   scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
-  geom_text(aes(x = 1993, y = 28), label = "First birth", color = "black", hjust = 0) +
-  geom_text(aes(x = 1993, y = 23), label = "Second birth", color = "black", hjust = 0) +
-  geom_text(aes(x = 1993, y = 12.5), label = "Third Birth", color = "black", hjust = 0) +
-  geom_text(aes(x = 1993, y = 8.5), label = "Fourth+ birth", color = "black", hjust = 0) +
+  geom_text(aes(x = 1993, y = 28), label = "First birth", color = "#f8766d", hjust = 0) +
+  geom_text(aes(x = 1993, y = 23), label = "Second birth", color = "#00bfc4", hjust = 0) +
+  geom_text(aes(x = 1993, y = 12.5), label = "Third Birth", color = "#c77cff", hjust = 0) +
+  geom_text(aes(x = 1993, y = 8.5), label = "Fourth+ birth", color = "#7cae00", hjust = 0) +
   theme_classic() +

### 2024-02-14 scripts/99-replications.R
update replication script to match quarto doc

@@ -47,12 +47,12 @@ fig_2a_data_long %>% ggplot(aes(x = year, y = Birth_Rate, color = Age_Group)) +
   labs(x = "Year", y = "Births per 1,000 women in \n relevant population subgroup", color = "Age Group") +
   scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +
   scale_x_continuous(limits = c(1980, 2020), breaks = seq(1980, 2020, by = 5)) +
-  geom_text(aes(x = 1985, y = 120), label = "Age 25-29", color = "#1e7837") +
-  geom_text(aes(x = 1985, y = 100), label = "Age 20-24", color = "#786d1e") +
-  geom_text(aes(x = 1985, y = 80), label = "Age 30-34", color = "#27a39b") +
+  geom_text(aes(x = 1985, y = 120), label = "Age 25-29", color = "#00ba38") +
+  geom_text(aes(x = 1985, y = 100), label = "Age 20-24", color = "#b79f00") +
+  geom_text(aes(x = 1985, y = 80), label = "Age 30-34", color = "#00bfc4") +
   geom_text(aes(x = 1985, y = 58), label = "Age 15-19", color = "#f8766d") +
-  geom_text(aes(x = 1985, y = 33), label = "Age 35-39", color = "#5587cf") +
-  geom_text(aes(x = 1985, y = 10), label = "Age 40-44", color = "#d45dce") +
+  geom_text(aes(x = 1985, y = 33), label = "Age 35-39", color = "#619cff") +
+  geom_text(aes(x = 1985, y = 10), label = "Age 40-44", color = "#f564e3") +
   theme_classic() +
   theme(legend.position = "none")
 
@@ -74,11 +74,11 @@ fig_2d_data_long %>% ggplot(aes(x = year, y = Birth_Rate, color = Education_Leve
   labs(x = "Year", y = "Births per 1,000 women in \n relevant population subgroup", color = "Education Level") +
   scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +
   scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
-  geom_text(aes(x = 1993, y = 112), label = "No high school degree", color = "#668a38", hjust = 0) +
+  geom_text(aes(x = 1993, y = 112), label = "No high school degree", color = "#7cae00", hjust = 0) +
[…]

### 2024-02-14 paper/paper.qmd
fixing an issue with my environment
added results for figs 1, 2a, 2b

@@
 
-## Data Visualization
+# Results
+
+@fig-1 showcases the trajectory of US birth rates spanning from 1980 to 2020, focusing on births per 1000 women aged 15 to 44.
+The peak occurred in 1990, reaching 70.9 births per 1000 women, closely followed by 2007 with 69.3 births per 1000 women.
+In contrast, 2020 marked the lowest point with only 55.8 births per 1000 women.
+From 1980 to 2007, the birth rate experienced moderate fluctuations, with a maximum variance of 7.3 births per 1000 women.
+However, following the year 2007, the birth rate seems to display a consistent a downward trend, declining by 13.5 births per 1000 women between 2007 and 2020.
 
@@
 ```
+
+Examining the trends in birth rates by five-year age groups reveals intriguing patterns.
+
+In the 15 - 19 years old age group, there has been a consistent decline in birth rates, punctuated by a brief period of growth from 1986 to 1991, peaking at 61.8 births per 1000 women in 1991 before steadily declining to its lowest recorded rate of 15.3 births per 1000 women in 2020.
+
+The 20-24 years old age group remained relatively stable from 1980 to 2007, reaching a peak of 116.5 births per 1000 women in 1990.
+However, it has experienced a consistent decline from 2007 onward, reaching a low of 62.8 births per 1000 women in 2020.
+
+Similarly, the 25-29 years old age group displayed stability until 2007, with the highest birth rate of 120.2 births per 1000 women recorded in 1990.
+Despite a gradual decline post-2007, it retained a relatively higher birth rate compared to younger age groups, hitting a low of 90.0 births per 1000 women in 2020.
+
+In contrast, the 30-34 years old age group witnessed a steady increase in birth rates over time, starting from 61.9 births per 1000 women in 1980 and peaking at 102.7 births per 1000 women in 2016.
+
+The 35-39 years old age group also experienced steady growth, with birth rates ranging from a low of 40.5 births per 1000 women in 1980 to a high of 52.8 births in 2019.
+
+The 40-44 years old age group displayed minimal change, remaining relatively steady over the years, with slight increases noted annually.
+The lowest birth rate was recorded in 1980 at 8.1 births per 1000 women, rising to 12.0 births per 1000 women in 2019.
+
+Analyzing the data in @fig-2a, the 25 - 29 years old age group consistently led in birth rates until being surpassed by the 30 - 34 years old age group in 2016.
+Conversely, the 40-44 years old age group consistently exhibited the lowest birth rates.
+However, with declining births in the 15-19 years old age group and a stable, increasing trend in the 40-44 years old age group, a shift in the age group with the lowest birth rate may soon occur for the first time in 40 years.
+
+Additionally, a noteworthy trend is the steady decline in birth rates among younger age groups from around 2007 onwards, while the older age groups have experienced the opposite trend.
 
@@
 
+Unlike the figures discussed previously, @fig-2b examines US birth rate trends concerning the mother's level of education, focusing on the period from 1992 to 2019 and women aged 20 to 44.
+
+The subgroup with the highest birth rate consistently comprises mothers without a high school degree.
+Peaking in 2006 at 125.0 births per 1000 women, this subgroup witnessed its lowest birth rate of 89.0 in 1996.
+Notably, it also exhibited the most fluctuations, maintaining relative stability from 1992 to 2005 with a slight downturn from 1994 to 1997.
+Then, it experienced rapid growth in 2006, sustaining similar birth rates over the following three years before declining.
+From 2010 to 2019, the birth rate returned to levels observed from 1992 to 2005.
+
+Mothers who have graduated high school are the sub-group with the most stable birth rate.
+This sub-group experienced its highest birth rate in 2008, with 77.4 births per 1000 women.
+Conversely, the lowest birth rate of 68.0 births occurred in 1997.
+
+Similar to the previous sub-group, mothers with some college education have maintained a relatively stable birth rate, while showing a slightly positive trend over the years.
+The lowest birth rate can be attributed to the year 1994, with 49.4 births per 1000 women, and the highest birth rate was recorded in 2015, at 65.5 births.
+This is also the sub-group that has historically had the lowest birth average rate.
+
+Lastly, the college graduates sub-group has also remained stable, showcasing a slight declining trend from 2016 onward.
+This sub-group's birth rate peaked in 2000 76.7 births per 1000 women.
+The lowest recorded birthrate was 59.4 in 2019.
 ```{r graph2b, fig.cap="Trends in Birth Rates by Mother's Level of Education (ages 20-44)"}

### 2024-02-14 paper/paper.qmd
fig 4 result section

@@
 
-@fig-1 showcases the trajectory of US birth rates spanning from 1980 to 2020, focusing on births per 1000 women aged 15 to 44.
+@fig-1 is the replication of Figure 1 from `The Puzzle of Falling US Birth Rates since the Great Recession' by @citeOriginalReplication, and showcases the trajectory of US birth rates spanning from 1980 to 2020, focusing on births per 1000 women aged 15 to 44.
 The peak occurred in 1990, reaching 70.9 births per 1000 women, closely followed by 2007 with 69.3 births per 1000 women.
@@
 
-Examining the trends in birth rates by five-year age groups reveals intriguing patterns.
+@fig-2a is the replication of Figure 2 (a) from the original paper. Examining the trends in birth rates by five-year age groups reveals intriguing patterns.
 
@@
 
-Unlike the figures discussed previously, @fig-2b examines US birth rate trends concerning the mother's level of education, focusing on the period from 1992 to 2019 and women aged 20 to 44.
+@fig-2b is the replication of Figure 2 (d) from the original paper, and examines US birth rate trends concerning the mother's level of education, focusing on the period from 1992 to 2019 and women aged 20 to 44.
 
@@
 ```
+
+@fig-2c replicates figure 2 (f) from @citeOriginalReplication, and breaks down the trends in birth rates by parity.
+
+In this case, all subgroups representing first, second, third, and fourth or more births exhibit nearly identical patterns, maintaining a stable birth rate.
+However, there is a notable distinction: the fourth or more birth subgroup shows slight growth from around 1995 to 2019, while all other parity groups witness a decline in births per 1000 women from approximately 2007 to 2019.
+
+Consistently, the first birth subgroup boasts the highest number of births per 1000 women, followed by the second birth subgroup.
+Conversely, the fourth or more birth subgroup, despite displaying a positive trend from 1995 to 2000, consistently records the lowest number of births per 1000 women each year, followed by the third birth subgroup.
 

### 2024-02-14 paper/paper.qmd
added discussion section to paper

@@
 ```
+# Results
+# Discussion
+
+## Leave the World Behind: A Shift in Priorities and the Evolving Social Norms
+
+The most likely explanation that acts as the root cause for the decline in US births is the shift in priorities across cohorts of women [@citeEconomicPolicy].
+While there are many who can speculate and point fingers at specific events that occurred pre- or post-2007 such as the "inevitable" financial crisis more commonly known as "The Great Recession".
+It is evident that there exists a relationship between the current state of the economy and fertility rate, in other words, fertility is procyclical [@citeSageJournals].
+This alone would suffice as the reasoning for such drop in fertility but instead the shifting priorities of young adults adopting a new and modern take on life is the overarching explanation.
+
+The social norm defines the boundaries or limits of how you should live your life.
+Walk the line of the social norm and you should expect to live the most ideal life, or so they say.
+The recent generations are an example of societies attempt at keeping pace with technological progression and redefining a social norm while challenging the bigger question at hand: "What is the meaning of life?" Should everyone have a similar goal in life?
+Does there exist a single task in the bucket list that every human shares?
+These are questions that young adults have been asking while being raised in a constantly evolving social landscape.
+Concerning parenthood, when men and women are ask the question "How does ones life  change when you have children", the most common answer is that your life becomes restricted now that you have to take on a new responsibility as a parent or idea that you have to giving up your freedom [@citePopulationCouncil].
+While this goal of this paper is not to convince to not have children, instead it aims to convey the perspective of young adults who have the choice of whether or not they wish to become parents of the next generation.
+
+DINK is an acronym for "dual income, no kids" and refers to couples who choose to not have children.
+While not being too unpopular of an ideology, is such decision considered selfish through the lens of the social norm?
+Nearly 48% of DINKs are spending the money that they are "saving" for not having any children towards travel.
+33% say that they enjoy the financial freedom of not having children.
+With 32% of DINK couples stating that the cost of living today is too high to have children and 26% say they cannot financially support a child right now.
+A high 68% of DINK couples do not plan to ever have children and do not expect to regret such decision [@citeDINK].
+With these statistics, can you really blame them?
+Regardless of whether or not these values become normalized into society, the declining birth rates in the US are a product of the shift in priorities for young adults.
+
 # References

### 2024-02-14 paper/paper.qmd
added minor tweaks

@@
 It is evident that there exists a relationship between the current state of the economy and fertility rate, in other words, fertility is procyclical [@citeSageJournals].
-This alone would suffice as the reasoning for such drop in fertility but instead the shifting priorities of young adults adopting a new and modern take on life is the overarching explanation.
+This alone would suffice as the reasoning for such drop in fertility but instead the shifting priorities of young adults adopting a new and modern take on life is the overarching explanation to the downward trajectory of US birth rates.
 

### 2024-02-14 scripts/00-simulate_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
-# Purpose: Simulates the data required to reproduce graphs
+# Purpose: Simulates the data that will be required to reproduce the figures
 # Author: Aaron Xiaozhou Liu, Jimmy Luc, Maroosh Gillani
-# Date: 13 February 2024
+# Date: 14 February 2024
 # Contact: maroosh.gillani@mail.utoronto.ca, aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca
 # License: MIT
 

### 2024-02-14 scripts/01-test_data.R

@@ -1,11 +1,11 @@
 #### Preamble ####
-# Purpose: Tests data acquired from original AEA paper's replication package
+# Purpose: Tests data acquired from original AEA article replication package
 # Author: Aaron Xiaozhou Liu, Jimmy Luc, Maroosh Gillani
-# Date: 13 February 2024
+# Date: 14 February 2024
 # Contact: maroosh.gillani@mail.utoronto.ca, aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: Must have data downloaded
-# Additional Information: these tests are applicable to the simulated data in 00-simulate_data.R
+# Pre-requisites: Must have the data downloaded
+# Additional Information: These tests are applicable to the simulated data in 00-simulate_data.R
 
 
 #### Workspace setup ####

### 2024-02-14 scripts/99-replications.R

@@ -1,10 +1,10 @@
 #### Preamble ####
 # Purpose: Replication of figures in the original paper
 # Author: Aaron Xiaozhou Liu, Jimmy Luc, Maroosh Gillani
-# Date: 12 February 2024
+# Date: 14 February 2024
 # Contact: maroosh.gillani@mail.utoronto.ca, aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: Must have data downloaded
+# Pre-requisites: Must have the data downloaded
 
 
 #### Workspace setup ####

### 2024-02-14 paper/paper.qmd
update paper abstract

@@
 We discover how different social, cultural, and economic shifts play a part in this worrying trend.
-This paper demonstrates how socioeconomic trends in the recent years have affected Americans, and how these patterns represented through data have played a key role in the irreversible downward trajectory of US birth rates."
+This paper demonstrates how socioeconomic trends in the recent years have affected Americans to shift their priorties, and how this change represented through data has played a key role in the irreversible downward trajectory of US birth rates."
 format:
@@
 ```
-# Results
+
 # Discussion
@@
 
+## Ethical Concerns and Biases
+
+## Weaknesses
+
+## Next Steps
+
 # References

### 2024-02-14 paper/paper.qmd
minor tweaks to paper

@@
 The peak occurred in 1990, reaching 70.9 births per 1000 women, closely followed by 2007 with 69.3 births per 1000 women.
-In contrast, 2020 marked the lowest point with only 55.8 births per 1000 women.
+In contrast, 2020 marked the lowest point with only 55.8 births per 1000 women which is nearly a 21.3% decrease in births since 1990.
 From 1980 to 2007, the birth rate experienced moderate fluctuations, with a maximum variance of 7.3 births per 1000 women.
-However, following the year 2007, the birth rate seems to display a consistent a downward trend, declining by 13.5 births per 1000 women between 2007 and 2020.
+However, following the year 2007, the birth rate seems to display a consistent a downward trend, declining by 13.5 births per 1000 women or a 19.5% decrease between 2007 and 2020.
 
@@
 
-@fig-2a is the replication of Figure 2 (a) from the original paper. Examining the trends in birth rates by five-year age groups reveals intriguing patterns.
+Next, @fig-2a is the replication of Figure 2 (a) from the original paper. Concerning itself with examining the trends in birth rates by five-year age groups, we can observe insightful results.
 
-In the 15 - 19 years old age group, there has been a consistent decline in birth rates, punctuated by a brief period of growth from 1986 to 1991, peaking at 61.8 births per 1000 women in 1991 before steadily declining to its lowest recorded rate of 15.3 births per 1000 women in 2020.
+Looking at the 15 - 19 years old age group, there has been a consistent decline in birth rates, punctuated by a brief period of growth from 1986 to 1991, peaking at 61.8 births per 1000 women in 1991 before steadily declining to its lowest recorded rate of 15.3 births per 1000 women in 2020.
 
 The 20-24 years old age group remained relatively stable from 1980 to 2007, reaching a peak of 116.5 births per 1000 women in 1990.
-However, it has experienced a consistent decline from 2007 onward, reaching a low of 62.8 births per 1000 women in 2020.
+However, similarly to the previous age group, it experiences a consistent decline from 2007 onward, reaching a low of 62.8 births per 1000 women in 2020.
 
-Similarly, the 25-29 years old age group displayed stability until 2007, with the highest birth rate of 120.2 births per 1000 women recorded in 1990.
+Again, the 25-29 years old age group displays stability until 2007, with the highest birth rate of 120.2 births per 1000 women recorded in 1990.
 Despite a gradual decline post-2007, it retained a relatively higher birth rate compared to younger age groups, hitting a low of 90.0 births per 1000 women in 2020.
@@
 
-The 40-44 years old age group displayed minimal change, remaining relatively steady over the years, with slight increases noted annually.
+Lastly, the 40-44 years old age group displayed minimal change, remaining relatively steady over the years, with slight increases noted annually.
 The lowest birth rate was recorded in 1980 at 8.1 births per 1000 women, rising to 12.0 births per 1000 women in 2019.
 
-Analyzing the data in @fig-2a, the 25 - 29 years old age group consistently led in birth rates until being surpassed by the 30 - 34 years old age group in 2016.
+Observing the data in @fig-2a, we find the 25 - 29 years old age group consistently led in birth rates until being surpassed by the 30 - 34 years old age group in 2016.
 Conversely, the 40-44 years old age group consistently exhibited the lowest birth rates.
@@
 
-Additionally, a noteworthy trend is the steady decline in birth rates among younger age groups from around 2007 onwards, while the older age groups have experienced the opposite trend.
+Additionally, a noteworthy trend is the steady decline in birth rates among younger age groups from around 2007 onward, while the older age groups have experienced the opposite trend.
 

### 2024-02-14 paper/paper.qmd
Updated variables of interest section

@@
 ## Variables of Interest
-Although the data contains an abundance of information, we are mainly concerned of US birth rates per year, the total female population, and various other factors including age, mother's level of education, and birth parity (number of children for a given women).
+Although the data contains an abundance of information, we are mainly concerned of US birth rates per year, the total female population, and various other factors including age, mother's level of education, and birth parity.
 As this paper focuses on birth rates in the US over the course of time, we have all the necessary and sufficient information required to proceed processing the data for our analysis.
 
-## Data Processing
+The over arching variable in our measurements is birth rate.
+Birth rate is measured by number of births per 1000 women.
+Using this metric, we graph it in the context of the other variables included in our research we stated above.
+
+Measuring birth rates by age allows us to split the data among different age groups of women.
+In doing so, we can observe how age affects the decision to have babies among women and how it might have changed over time.
+Another metric we will be comparing is the mother's level of education.
+Women from different education backgrounds would have impacts on fertility rates, and this measurement allows us to observe the differences between these education backgrounds.
+Lastly, we will be graphing birth parity, which is the measurement of how many children a given woman has.
+Birth parity would allows us to see the rates of which women have one child, two children and so on, and compare the frequencies of each to each other and over the course of the past few decades.
+
+## Data Measurement
 

### 2024-02-14 paper/paper.qmd
added doi link to paper

@@
 thanks: "Code and data are available at: https://github.com/rex009x/us_birth_rates/tree/master/data.
-Selected aspects of the original paper that underpins this paper were reproduced and a record of this is available at: https://www.socialsciencereproduction.org/reproductions/bfe97499-4c05-466f-8cb5-fbedc27a83a1/index "
+Selected aspects of the original paper that underpins this paper were reproduced and a record of this is available at: https://doi.org/10.48152/ssrp-nsc2-4v21 "
 date: today

### 2024-02-14 paper/paper.qmd
Updated title

@@
 ---
-title: "From Trends to Truth: A Comprehensive Analysis Revealing Contributing Factors Behind Declining US Birth Rates"
+title: "A Comprehensive Analysis Revealing Contributing Factors Behind Declining US Birth Rates"
 #subtitle: "My subtitle if needed"

### 2024-02-15 paper/paper.qmd
added remaining subsections of paper

@@
 
-## Data Measurement
+## Data Processing
+
+The data we acquired through various sources and methods as previously mentioned presents a plethora of data points which we concern ourselves with.
+In order to produce our desired visualizations, we must first begin by process our data in preparation.
+While a portion of the data has already been processed by the authors of "The Puzzle of Falling US Birth Rates since the Great Recession" [@citeOriginalReplication], the remaining data sets are abundant.
+Data that can be so specific such as the population of women the age of 15 who live in Arkansas.
+
+With the consideration of our variables of interest in mind, we can proceed with the processing of the data.
+As birth rates are the main focal point of this paper, we are interested in the number of births that occur throughout the United States and the total population at any given time frame.
+The data including US Census Data and NCHS Vital Statistics Natality Birth Data provides this information by listing out the number of women and child births per state by year.
+Together, we are have the sufficient information required to proceed with the visualization for the number of births per 1000 women in the US.
+
+Next, we are interested to find the number of births per 1000 women in regards to the woman's age.
+As previously mentioned, we have the required data points concerning the population of women but now incorporating another variable of interest being the age.
+Proceeding, we intend to organize the data into age groups in order to observe the potential generational trends in birth rate.
+Fortunately, the data is already prepared containing the birth rates per year for each corresponding five-year age group [@citeOriginalReplication].
+
+Moving onto our next data point of interest, the mother's level of education requires the collection of data that records the population of women according to their levels of education.
+Utilizing the Current Population Survey (CPS) organized by the U.S.
+Census Bureau, the data is prepared containing the birth rates for each level of education by year respectively [@citeOriginalReplication].
+
+Considering births by "parity", referring to the number of children for a given women, we are interested in the data containing information regarding the number of first, second, third, and fourth+ births in the US.
+Combining NCHS Natality Microdata and CDC SEER population data, and simply filter for the data points of interest to receive the information required to generate our visualizations.
+
+Similarly to the data processing to acquire the birth rates per year by five-year age groups, gathering NCHS Vital Statistics birth data and CDC SEER population data we organize the data by mother's age and birth cohorts.
+Data processing and visualization of @fig-3 was completed with the aid of Kimlin Chin's replication of Kearney, Levine, and Pardue's (2020) article "The Puzzle of Falling US Birth Rates since the Great Recession".
+
+## Measurement
+
+The measurement of birth rates is nothing short of a difficult task upon consideration of all its nuances.
+What exactly defines a birth and what parameters determine how it is recorded?
+This variable of interest becomes slightly cloud when answering these questions.
+Do births that occur outside of a hospital count to the total number of births?
+The bounds of what a birth is becomes even blurrier when we consider miscarriages and related factors.
+
+Age being data of interest proves to be one of the most reliable and sound measurement throughout this analysis.
+Data collected through government initiated surveys regardless of whether or not it is mandatory, requires proper documented identification.
+While the exact date of birth remains confidential, the respondent's year of birth or age suffices for the purposes of this analysis.
+
+Measuring the level of education appears to be simple but is more inconsistent than initial thought.
+Data acquired through the CPS which is a monthly voluntary survey where U.S.
+households are selected [@citeCPS].
+This is important to note as there may be traces of response bias and other forms of biases.
+As a result, it is difficult to filter the data from these bias entries and the generated visualizations may be skewed.
+
+The number of children for a given mother or "parity" is a rather straight-forward measurement.
+Although questions concerning whether legal adoption of a child contributes to this measurement, it is unclear to find an answer.
+A more concise explanation for the properties of this measurement would greatly benefit the readers experience.
 
@@
 
+Lastly, Figure 5 from the original paper is replicated in @fig-3, and showcases children ever born based on the mother's age and birth cohort.
+
+All cohorts exhibit a consistent pattern of increasing children ever born as the mother ages.
+In particular, the first three spanning 1968-1972, 1973-1977, and 1978-1982 are nearly identical, especially from the ages of 15 to 26.
+After the mother reaches an age of 26 years old, there are only marginal differences among these cohorts.
+
+In contrast, while the latter three cohorts still generally follow the same trend of an increasing number of children ever born as the mother ages, it must be noted that the children ever born are steadily declining with each subsequent cohort.
+Thus, the most recent cohort of 1993-1997 records the lowest number of children ever born among all six cohorts.
+There is also a more significant variance between children ever born after the mother reaches an age of approximately 17 years old in the three most recent cohorts.
+
 ```{r graph3, fig.cap="Children Ever Born by Mother's Age and Birth Cohort"}
@@
 
-## Ethical Concerns and Biases
+## Weaknesses
 
-## Weaknesses
+While we were able to find very interesting trends and connections with our data, there are still a few weaknesses within our data.
+Some ethical concerns and biases are likely present as the data is based on a survey of women.
+One of these concerns is that it does not take into the account those who identify as LGBTQ+.
+The data we used in our analysis range from 1980 to 2020, but it wasn't until 2021 that the options to identify as LGBTQ+ was added to the census [@citeCensus].
+Due to these options not being included, it could lead to these people classifying themselves differently.
+Since the data is based on those who identify as female, it's likely that there are those are pregnant to not be counted.
+
+Another weakness within the data is from the education graph.
+In the graphs, the options are no high school degree, high school graduate, some college, and college graduate.
+The data could potentially be slightly skewed by the "some college" section, as this could include students, but also those who did not complete college and dropped out, in which case it would be more appropriate to group them in the "High school graduate" section.
+Another weakness with this data is the lack of any options beyond college graduate.
+People that may have completed masters or PHD programs are also included in this section, when they could be their own separate sections, or a new section that accounts for post graduate programs.
 
@@
 
+Some next steps that can be taken is the inclusion of those who identify in minority groups.
+This would include their inputs in the graphs, and instead of birth rate being measured by births per 1000 women, it would be birth per 1000 people to account for the minority groups.
+Other steps that can be taken is to consider other factors, such as ethnic backgrounds, seeing which demographics have a higher birth rates.
+Another variable of interest to be considered in the future is if people are immigrants, as people from those groups are potentially not documented in the data.
+Lastly, an improvement to the education graph as stated before in the weakness, combine the "Some college" group with "High school graduate" and splitting up "College graduate" to account for those who have completed programs, such as a masters, after college.
+
+\newpage
+
 # References

### 2024-02-15 scripts/00-simulate_data.R
added tests for 00-simulate_data.R

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Simulates the data that will be required to reproduce the figures
+# Purpose: Simulates and tests the data that will be required to reproduce the figures
 # Author: Aaron Xiaozhou Liu, Jimmy Luc, Maroosh Gillani
 # Date: 14 February 2024
 # Contact: maroosh.gillani@mail.utoronto.ca, aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca
@@ -33,7 +33,7 @@ simulate_fig_2a <-
     brate_2529 = rnorm(fig_2a_years, mean = 110, sd = 5),
     brate_3034 = rnorm(fig_2a_years, mean = 90, sd = 10),
     brate_3539 = rnorm(fig_2a_years, mean = 40, sd = 10),
-    brate_4044 = rnorm(fig_2a_years, mean = 5, sd = 5)
+    brate_4044 = rnorm(fig_2a_years, mean = 5, sd = 2)
   )
 
 # simulate figure 2d data
@@ -66,3 +66,63 @@ simulate_fig_5 <-
     cohort2 = rep(1:6, each = 5, length.out = length(fig_5_motherage)),
     numbirth = rnorm(fig_5_motherage, mean = 700000, sd = 200000)
   )
+
+#### Test data ####
+## test data types##
+# methods used: https://stackoverflow.com/a/29591221
+all(sapply(simulate_fig_1$year, is.numeric))
+all(sapply(simulate_fig_1$brate_all, is.numeric))
+
+all(sapply(simulate_fig_2a$year, is.numeric))
+all(sapply(simulate_fig_2a$brate_1519, is.numeric))
+all(sapply(simulate_fig_2a$brate_2024, is.numeric))
+all(sapply(simulate_fig_2a$brate_2529, is.numeric))
+all(sapply(simulate_fig_2a$brate_3034, is.numeric))
+all(sapply(simulate_fig_2a$brate_3539, is.numeric))
+all(sapply(simulate_fig_2a$brate_4044, is.numeric))
+
+all(sapply(simulate_fig_2d$year, is.numeric))
+all(sapply(simulate_fig_2d$brate_hsdropout, is.numeric)) # is.numeric ignores NA values
+all(sapply(simulate_fig_2d$brate_hsgrad, is.numeric))
[…]

### 2024-02-15 scripts/01-test_data.R

@@ -1,12 +1,10 @@
 #### Preamble ####
-# Purpose: Tests data acquired from original AEA article replication package
+# Purpose: Tests data acquired from original paper replication package
 # Author: Aaron Xiaozhou Liu, Jimmy Luc, Maroosh Gillani
 # Date: 14 February 2024
 # Contact: maroosh.gillani@mail.utoronto.ca, aaronxiaozhou.liu@mail.utoronto.ca, jimmy.luc@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: Must have the data downloaded
-# Additional Information: These tests are applicable to the simulated data in 00-simulate_data.R
-
 
 #### Workspace setup ####
 library(tidyverse)

### 2024-02-15 paper/paper.qmd
added minor tweaks

@@
 date-format: long
-abstract: "Over the last couple of years, the American population has been witnessing a steady decline in their birth rates, as suggested by the American Economic Association's article 'The Puzzle of Falling US Birth Rates since the Great Recession' by Kearney, Levine, and Pardue (2020).
+abstract: "The United States has been experiencing a steady decline in their birth rates, as suggested by Kearney, Levine, and Pardue's (2020) article 'The Puzzle of Falling US Birth Rates since the Great Recession'.
 Through the analysis of multiple different factors, such as age and education, we notice a continual decline in birth rate starting from the year 2007.
 We discover how different social, cultural, and economic shifts play a part in this worrying trend.
-This paper demonstrates how socioeconomic trends in the recent years have affected Americans to shift their priorties, and how this change represented through data has played a key role in the irreversible downward trajectory of US birth rates."
+This paper demonstrates how socioeconomic trends have affected Americans to shift their priorities, and how this change represented through data has played a key role in the irreversible downward trajectory of US birth rates."
 format:
@@
 
-The nations birth rate has fallen significantly since 2007, dropping but about 23% by 2022, measured by number of births per 1000 women [@citeAxios] and that birth rates tend to fall as based on the economy.
+The nation's birth rate has fallen significantly since 2007, dropping about 23% by 2022, measured by the number of births per 1000 women [@citeAxios] and that birth rates tend to fall based on the economy.
 2007 marks the start of the most severe economic crisis the world has ever faced, spanning the years 2007 until 2009.
-The events of this crisis has had long lasting damage on the economy, with unemployment doubling and the crash of the housing market [@citeFDR].
+The events of this crisis hve had long lasting damage on the economy, with unemployment doubling and the crash of the housing market [@citeFDR].
 These poor economic conditions and the unusually slow recovery of the economy would play a large role in the decision to have children, as an unpredictable economic environment makes people feel less secure about raising a family, thus contributing to the recorded decline in birth rates.
@@
 
-The datasets for this paper were acquired through the American Economic Association's paper "The Puzzle of Falling US Birth Rates since the Great Recession" [@citeOriginalReplication], with the aid of data available from Centers for Disease Control, and Prevention (CDC) Natality data [@citeCDCNCHS] and CDC SEER U.S.
+The data for this paper was acquired through the replication package of Kearney, Levine, and Pardue's (2020) article "The Puzzle of Falling US Birth Rates since the Great Recession" available through the American Economic Association (AEA), with the aid of data available from Centers for Disease Control, and Prevention (CDC) Natality data [@citeCDCNCHS] and CDC SEER U.S.
 Population Data [@citeCDCSeer].
@@
 The United States census is legally mandated by the United States Constitution which occurs every 10 years.
-The data is largely comprised of `numeric` values representing years, birth rates by race and ethnicity, marital status, education, age, and birth parity, or `character` values representing US states.
+The data consists largely of `numeric` values representing years, birth rates by race and ethnicity, marital status, education, age, and birth parity, or `character` values representing US states.
 
@@
 ## Variables of Interest
-Although the data contains an abundance of information, we are mainly concerned of US birth rates per year, the total female population, and various other factors including age, mother's level of education, and birth parity.
+Although the data contains an abundance of information, we are mainly concerned with US birth rates per year, the total female population, and various other factors including age, mother's level of education, and birth parity.
 As this paper focuses on birth rates in the US over the course of time, we have all the necessary and sufficient information required to proceed processing the data for our analysis.
 
-The over arching variable in our measurements is birth rate.
-Birth rate is measured by number of births per 1000 women.
+The overarching variable in our measurements is birth rate.
+Birth rate is measured by the number of births per 1000 women.
 Using this metric, we graph it in the context of the other variables included in our research we stated above.
@@
 Lastly, we will be graphing birth parity, which is the measurement of how many children a given woman has.
-Birth parity would allows us to see the rates of which women have one child, two children and so on, and compare the frequencies of each to each other and over the course of the past few decades.
+Birth parity would allow us to see the rates of which women have one child, two children and so on, and compare the frequencies of each to each other and over the course of the past few decades.
 
@@
 The data we acquired through various sources and methods as previously mentioned presents a plethora of data points which we concern ourselves with.
-In order to produce our desired visualizations, we must first begin by process our data in preparation.
+In order to produce our desired visualizations, we must first begin by processing our data in preparation.
 While a portion of the data has already been processed by the authors of "The Puzzle of Falling US Birth Rates since the Great Recession" [@citeOriginalReplication], the remaining data sets are abundant.
@@
 The data including US Census Data and NCHS Vital Statistics Natality Birth Data provides this information by listing out the number of women and child births per state by year.
-Together, we are have the sufficient information required to proceed with the visualization for the number of births per 1000 women in the US.
+Together, we have the sufficient information required to proceed with the visualization for the number of births per 1000 women in the US.
 
@@
 
-Considering births by "parity", referring to the number of children for a given women, we are interested in the data containing information regarding the number of first, second, third, and fourth+ births in the US.
+Considering births by "parity", referring to the number of children for a given woman, we are interested in the data containing information regarding the number of first, second, third, and fourth+ births in the US.
 Combining NCHS Natality Microdata and CDC SEER population data, and simply filter for the data points of interest to receive the information required to generate our visualizations.
@@
 What exactly defines a birth and what parameters determine how it is recorded?
-This variable of interest becomes slightly cloud when answering these questions.
+This variable of interest becomes slightly cloudy when answering these questions.
 Do births that occur outside of a hospital count to the total number of births?
@@
 
-Age being data of interest proves to be one of the most reliable and sound measurement throughout this analysis.
+Age being data of interest proves to be one of the most reliable and sound measurements throughout this analysis.
 Data collected through government initiated surveys regardless of whether or not it is mandatory, requires proper documented identification.
@@
 Although questions concerning whether legal adoption of a child contributes to this measurement, it is unclear to find an answer.
-A more concise explanation for the properties of this measurement would greatly benefit the readers experience.
+A more concise explanation for the properties of this measurement would greatly benefit the reader's experience.
 
@@
 While there are many who can speculate and point fingers at specific events that occurred pre- or post-2007 such as the "inevitable" financial crisis more commonly known as "The Great Recession".
-It is evident that there exists a relationship between the current state of the economy and fertility rate, in other words, fertility is procyclical [@citeSageJournals].
+Fertility is procyclical: there exists a relationship between the current economic state and fertility rate [@citeSageJournals].
 This alone would suffice as the reasoning for such drop in fertility but instead the shifting priorities of young adults adopting a new and modern take on life is the overarching explanation to the downward trajectory of US birth rates.

### 2024-02-15 scripts/00-simulate_data.R

@@ -12,7 +12,7 @@ library(tidyverse)
 # set seed for simulation
 set.seed(416)
 
-# simulate figure 1 data
+# simulate figure 1 data - Trend in US birth rates from 1980 to 2020
 # methods used: https://stackoverflow.com/a/68763213
 fig_1_years = 1980:2020
 fig_1_brate_mean = 65
@@ -23,7 +23,7 @@ simulate_fig_1 <-
     brate_all = rnorm(n = fig_1_years, mean = fig_1_brate_mean, sd = fig_1_brate_sd)
     )
 
-# simulate figure 2a data
+# simulate figure 2a data - Trend in birth rates by five-year age group
 fig_2a_years = 1980:2020
 simulate_fig_2a <-
   tibble(
@@ -36,7 +36,7 @@ simulate_fig_2a <-
     brate_4044 = rnorm(fig_2a_years, mean = 5, sd = 2)
   )
 
-# simulate figure 2d data
+# simulate figure 2d data - Trend in birth rate by mother's level of education
 fig_2d_years = 1990:2020
 simulate_fig_2d <-
   tibble(
@@ -47,7 +47,7 @@ simulate_fig_2d <-
     brate_colgrad = rnorm(fig_2d_years, mean = 70, sd = 5)
   )
 
-# simulate figure 2f data
+# simulate figure 2f data - Trend in birth rates by parity (number of children for a given woman)
 fig_2f_years = 1990:2020
 simulate_fig_2f <-
   tibble(
@@ -58,7 +58,7 @@ simulate_fig_2f <-
     numbirth_fourthbirth = rnorm(fig_2f_years, mean = 9, sd = 0.5)
   )
 
-# simulate figure 5 data
+# simulate figure 5 data - Number of children ever born by mother's age and birth cohort
 fig_5_motherage = 15:44
 simulate_fig_5 <-
   tibble(

### 2024-02-15 paper/paper.qmd
update final tweaks

@@
 Similarly to the data processing to acquire the birth rates per year by five-year age groups, gathering NCHS Vital Statistics birth data and CDC SEER population data we organize the data by mother's age and birth cohorts.
-Data processing and visualization of @fig-3 was completed with the aid of Kimlin Chin's replication of Kearney, Levine, and Pardue's (2020) article "The Puzzle of Falling US Birth Rates since the Great Recession".
+Data processing and visualization of @fig-3 was completed with the aid of Kimlin Chin's replication [@citeAnotherReplication] of Kearney, Levine, and Pardue's (2020) article "The Puzzle of Falling US Birth Rates since the Great Recession".
 

### 2024-02-15 paper/paper.qmd
fixes and tweaks and I just wanna sleep

@@
 We discover how different social, cultural, and economic shifts play a part in this worrying trend.
-This paper demonstrates how socioeconomic trends have affected Americans to shift their priorities, and how this change represented through data has played a key role in the irreversible downward trajectory of US birth rates."
+This paper demonstrates how socioeconomic trends have affected Americans to shift their priorities, and how this change represented through data has played a key role in the downward trajectory of US birth rates."
 format:
@@
 
-The nation's birth rate has fallen significantly since 2007, dropping about 23% by 2022, measured by the number of births per 1000 women [@citeAxios] and that birth rates tend to fall based on the economy.
-2007 marks the start of the most severe economic crisis the world has ever faced, spanning the years 2007 until 2009.
-The events of this crisis hve had long lasting damage on the economy, with unemployment doubling and the crash of the housing market [@citeFDR].
+The nation's birth rate has fallen significantly since 2007, dropping about 23% by 2022, as measured by the number of births per 1000 women [@citeAxios], and birth rates tend to fall based on the economy.
+2007 marks the start of the most severe economic crisis the world has ever faced, spanning until 2009.
+The events of this crisis have had long lasting damage on the economy, with unemployment doubling and the crash of the housing market [@citeFDR].
 These poor economic conditions and the unusually slow recovery of the economy would play a large role in the decision to have children, as an unpredictable economic environment makes people feel less secure about raising a family, thus contributing to the recorded decline in birth rates.
@@
 Population Data [@citeCDCSeer].
-The data sources were accessed on February 12th, 2024 and acts as the foundation of this paper.
-The data sources themselves are a collaborative effort including but not limited to the U.S.
+The data sources were accessed on February 12th, 2024 and act as the foundation of this paper.
+The data sources themselves are a collaborative effort including, but not limited to, the U.S.
 Census Bureau, the National Cancer Institute, SEER Data, CDC, and the National Center for Health Statistics (NCHS).
@@
 
-All data collection and analysis was done using statistical tools and data visualization program R [@citeR] and Rstudio IDE [@citeRStudio] to help streamline the workflow.
+All data collection and analysis was done using statistical tools such as R [@citeR], and the data visualization program Rstudio IDE [@citeRStudio] to help streamline the workflow.
 The analysis was done using the R program and the following supporting packages `tidyverse` [@citeTidyverse], `haven` [@citeHaven], `ggplot2` [@citeGgplot2], `here` [@citeHere], `ggthemes` [@citeGgthemes], and `dplyr` [@citeDplyr].
@@
 Birth rate is measured by the number of births per 1000 women.
-Using this metric, we graph it in the context of the other variables included in our research we stated above.
+Using this metric, we graph it in the context of the other variables included in our research, as stated above.
 
@@
 
-The data we acquired through various sources and methods as previously mentioned presents a plethora of data points which we concern ourselves with.
+The data we acquired through various sources and methods, as previously mentioned, presents a plethora of data points which we concern ourselves with.
 In order to produce our desired visualizations, we must first begin by processing our data in preparation.
 While a portion of the data has already been processed by the authors of "The Puzzle of Falling US Birth Rates since the Great Recession" [@citeOriginalReplication], the remaining data sets are abundant.
-Data that can be so specific such as the population of women the age of 15 who live in Arkansas.
+For example, there is data that is very specific, such as the population of women the age of 15 who live in Arkansas.
 
-With the consideration of our variables of interest in mind, we can proceed with the processing of the data.
+With consideration of our variables of interest in mind, we can proceed with the processing of the data.
 As birth rates are the main focal point of this paper, we are interested in the number of births that occur throughout the United States and the total population at any given time frame.
 The data including US Census Data and NCHS Vital Statistics Natality Birth Data provides this information by listing out the number of women and child births per state by year.
-Together, we have the sufficient information required to proceed with the visualization for the number of births per 1000 women in the US.
+Together, this provides us with sufficient information required to proceed with the visualization for the number of births per 1000 women in the US.
 
-Next, we are interested to find the number of births per 1000 women in regards to the woman's age.
-As previously mentioned, we have the required data points concerning the population of women but now incorporating another variable of interest being the age.
-Proceeding, we intend to organize the data into age groups in order to observe the potential generational trends in birth rate.
+Next, we are interested in finding the number of births per 1000 women with regards to the woman's age.
+As previously mentioned, we have the required data points concerning the population of women, but now incorporating another variable of interest being the age.
+Then, we intend to organize the data into age groups in order to observe the potential generational trends in birth rate.
 Fortunately, the data is already prepared containing the birth rates per year for each corresponding five-year age group [@citeOriginalReplication].
@@
 Considering births by "parity", referring to the number of children for a given woman, we are interested in the data containing information regarding the number of first, second, third, and fourth+ births in the US.
-Combining NCHS Natality Microdata and CDC SEER population data, and simply filter for the data points of interest to receive the information required to generate our visualizations.
+Combining NCHS Natality Microdata and CDC SEER population data, we simply filter for the data points of interest to receive the information required to generate our visualizations.
 
-Similarly to the data processing to acquire the birth rates per year by five-year age groups, gathering NCHS Vital Statistics birth data and CDC SEER population data we organize the data by mother's age and birth cohorts.
+Similar to the data processing to acquire the birth rates per year by five-year age groups, we gather and organize NCHS Vital Statistics birth data and CDC SEER population data by mother's age and birth cohorts.
 Data processing and visualization of @fig-3 was completed with the aid of Kimlin Chin's replication [@citeAnotherReplication] of Kearney, Levine, and Pardue's (2020) article "The Puzzle of Falling US Birth Rates since the Great Recession".
@@
 
-The measurement of birth rates is nothing short of a difficult task upon consideration of all its nuances.
+The measurement of birth rates is nothing short of a difficult task considering all of its nuances.
 What exactly defines a birth and what parameters determine how it is recorded?
@@
 
-Measuring the level of education appears to be simple but is more inconsistent than initial thought.
-Data acquired through the CPS which is a monthly voluntary survey where U.S.
+Measuring the level of education appears to be simple but is more inconsistent than initially thought.
+Data is acquired through the CPS, which is a monthly voluntary survey where U.S.
 households are selected [@citeCPS].
 This is important to note as there may be traces of response bias and other forms of biases.
-As a result, it is difficult to filter the data from these bias entries and the generated visualizations may be skewed.
+As a result, it is difficult to filter the data from these biased entries and the generated visualizations may be skewed.
 
 The number of children for a given mother or "parity" is a rather straight-forward measurement.
-Although questions concerning whether legal adoption of a child contributes to this measurement, it is unclear to find an answer.
-A more concise explanation for the properties of this measurement would greatly benefit the reader's experience.
+Although questions concerning whether legal adoption of a child contributes to this measurement remain, it is difficult to find an answer.
+A more concise explanation for the properties of this measurement would greatly benefit the reader and replicator's experience.
 
@@
 The most likely explanation that acts as the root cause for the decline in US births is the shift in priorities across cohorts of women [@citeEconomicPolicy].
-While there are many who can speculate and point fingers at specific events that occurred pre- or post-2007 such as the "inevitable" financial crisis more commonly known as "The Great Recession".
-Fertility is procyclical: there exists a relationship between the current economic state and fertility rate [@citeSageJournals].
-This alone would suffice as the reasoning for such drop in fertility but instead the shifting priorities of young adults adopting a new and modern take on life is the overarching explanation to the downward trajectory of US birth rates.
+While there are many who can speculate and point fingers at specific events that occurred pre- or post-2007, such as the "inevitable" financial crisis more commonly known as "The Great Recession".
+In fact, @fig-1 implies a strong correlation between the 2007 crisis and declining birth rates, as it is clear to see that births per 1000 women have been on a sharp downwards trajectory since the aforementioned year.
+
+Another factor to consider is that Fertility is pro-cyclical: there exists a relationship between the current economic state and fertility rate [@citeSageJournals].
+This alone would suffice as the reasoning for such drop in fertility but instead the shifting priorities of young adults, adopting a new and modern take on life is the overarching explanation to the downward trajectory of US birth rates.
+This is supported by @fig-2a, as younger women, ranging from the ages of 15 to 29 years old have all exhibited decreasing births.
+@citeAgeArticle provides a possible explanation for this phenomenon.
+They state that younger women are more focused on their educations and careers, for the purpose of ensuring financial stability.
+Thus, they might feel unready to support a baby in their circumstances.
+In contrast, a comparatively older woman will have most likely become comfortable financially, and would feel ready to have, and support a baby.
+
+@citeAgeArticle's point regarding education brings up some doubts regarding results seen in @fig-2b.
+Women pursuing higher education share similar birth rates over the years, all of which are significantly lower than women who do not have a high school degree.
+On one hand, one could justify this by considering the fact that women pursuing education likely have lesser time in their day, but at the same time, there are questions about how lesser educated women are able to financially afford taking care of their babies.
+@citeEducationArticle helps clear some of these doubts by pointing out that educated women have better access to birth control.
+Furthermore, they are also able to provide better care for their babies, resulting in an increased human capital for the baby, subsequently reducing the need for more children [@citeEducationArticle].
+
+With @citeEducationArticle's information on why educated women have less children in mind, we look at the fact that female education in North America has experienced a steady rise from 1970 to 2020 [@citeNAFemEducation].
+This can actually be seen in @fig-2c, which shows that lesser women have more than one child.
+A similar trend is seen in @fig-3, where the number of children ever born are declining at a significant rate as they are born to women belonging to more recent cohorts.
+This could be another explanation as to why birth rates are declining overall in the United States.
 
